---
title: "Roybal - Efficacy"
format:
  html:
    toc: true
    embed-resources: true
editor: visual
editor_options:
  chunk_output_group: inline
---

# Roybal Final Data Analysis — Efficacy

**Purpose:** This document covers preliminary efficacy analyses of walking outcomes from the Roybal pilot RCT.

------------------------------------------------------------------------

### Load libraries

```{r}
pkgs <- c(
  "readr", "readxl", "tidyverse", "lme4", "lmerTest",
  "performance", "emmeans", "pwr", "longpower", "simr",
  "ggplot2", "kableExtra", "broom.mixed", "car", "gridExtra",
  "ggridges"
)

install.packages(setdiff(pkgs, rownames(installed.packages())))

invisible(lapply(pkgs, library, character.only = TRUE))
```

### Load data

```{r}
load("roybal_base.RData")

# Filter to weeks 1-12, join covariates, scale days since injury
weekly_walks <- weekly_walks %>%
  filter(week <= 12) %>%
  left_join(
    completed %>% select(id, age2, sex_cat, years_of_ed, days_since_injury),
    by = "id"
  ) %>%
  mutate(days_since_injury_scaled = scale(days_since_injury)[, 1])

# df and df_means — used for ribbon plots in Model 3 and Model 4 visualizations
df <- weekly_walks %>%
  rename(group = type, duration = total_duration_walks) %>%
  mutate(duration = if_else(id == "Roybal_60" & week == 1, NA_real_, duration))

df_means <- df %>%
  group_by(group, week) %>%
  summarise(
    across(
      c(weekly_walks, duration, total_steps_in_walks),
      list(mean = ~ mean(., na.rm = TRUE),
           se   = ~ sd(., na.rm = TRUE) / sqrt(sum(!is.na(.)))),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )
```

## Efficacy

### Model 1: Total Walks Over Time (Group × Week Interaction) - Model + Covariates - both walking criteria (2x+15 mins + \>27mins)

```{r}
# Adjusted Model (race excluded due to limited sample diversity)
glmodel1_adj <- glmer(weekly_walks ~ type * week + age2 + sex_cat +
                      years_of_ed + days_since_injury_scaled + (1 + week | id),
                      control = glmerControl(optimizer = "bobyqa"),
                      data = weekly_walks,
                      family = poisson(link = "log"))
summary(glmodel1_adj)

# Extract IRRs and CIs
tidy(glmodel1_adj, effects = "fixed") %>%
  mutate(
    IRR = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error)
  )


# Model comparison to check random inercept only vs intercept and slope models - Poisson model shows best fit
#AIC(glmodel1_adj_i, glmodel1_adj, glmodel1)
#BIC(glmodel1_adj_i, glmodel1_adj, glmodel1)
```

### Model 1: Post-hoc Power Analysis (Group × Week Interaction)

```{r}
# Post-hoc power analysis for the Group × Week interaction term
# Requires from model summary: estimate, std.error, and statistic (z-value)
# for the typeIntervention:week term

# --- Extract interaction term from fitted model ---
coef_summary <- tidy(glmodel1_adj, effects = "fixed")

interaction_term <- coef_summary %>%
  filter(term == "typeIntervention:week")

log_irr   <- interaction_term$estimate   # log-scale effect size (from Estimate column)
se        <- interaction_term$std.error  # std error (from Std. Error column)
z_stat    <- interaction_term$statistic  # z-value (from z value column)
irr       <- exp(log_irr)               # exponentiated = Incidence Rate Ratio per week

# --- Sample sizes per group ---
# Pull directly from the model frame (always available after fitting)
n_per_group <- glmodel1_adj@frame %>%
  distinct(id, type) %>%
  count(type, name = "n_participants")

n_total <- sum(n_per_group$n_participants)

# --- Method 1: Wald-based analytical approximation ---
# Achieved power = P(reject H0 | effect = observed effect, alpha = 0.05)
# For a two-sided Wald test: power ≈ Φ(|z| − z_{α/2})
alpha   <- 0.05
z_alpha <- qnorm(1 - alpha / 2)

power_wald <- pnorm(abs(z_stat) - z_alpha) +   # upper tail
              pnorm(-abs(z_stat) - z_alpha)      # lower tail (negligible but exact)

cat("=== Post-hoc Power Analysis: Group × Week Interaction ===\n\n")
cat("Values extracted from model summary:\n")
cat("  Estimate (log IRR per week):", round(log_irr, 4), "\n")
cat("  IRR per week               :", round(irr, 4), "\n")
cat("  Std. Error                 :", round(se, 4), "\n")
cat("  z-statistic                :", round(z_stat, 4), "\n\n")
cat("Sample sizes:\n")
print(n_per_group)
cat("  Total N                    :", n_total, "\n\n")
cat("Achieved power (Wald, alpha =", alpha, "):", round(power_wald, 4), "\n")

# --- Method 2: Simulation-based power via simr (more accurate, ~slow) ---
# Simulates data from the fitted model and re-fits nsim times to estimate
# the proportion of runs where the interaction p-value < alpha.
# Uncomment to run — nsim = 500 is reasonable; 1000 is more precise.

#set.seed(42)
#power_sim <- powerSim(glmodel1_adj,
                       test = fixed("typeIntervention:week", #method = "z"),
                       nsim = 500)
#print(power_sim)
```

### Model 1: Estimated marginal means at weeks 1, 6, 12 and exponentiated coefficients

```{r}
# Estimated marginal means from adjusted model
emm1_adj <- emmeans(glmodel1_adj, ~ type | week, at = list(week = c(1, 6, 12)), type = "response")
print(emm1_adj)

# Pairwise comparisons: Intervention vs Control at each week
pairs(emm1_adj, by = "week")

# table this for pub

# Extract EMMs
emm_table <- emm1_adj %>%
  as.data.frame() %>%
  select(week, type, rate) %>%
  pivot_wider(names_from = type, values_from = rate) %>%
  mutate(
    Ratio = Intervention / Control,
    Ratio_Label = paste0(round(Ratio, 1), "× more")
  )

print(emm_table)

emm_table_with_p <- emm_table %>%
  mutate(`P-value` = c("<0.001", "<0.001", "<0.001"))


# Create kable table
table_kable <- emm_table_with_p %>%
  kbl(caption = "Estimated Marginal Means of Weekly Walks",
      booktabs = TRUE,
      align = "c",
      digits = 2) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)

table_kable

# Save as PDF (requires LaTeX)
save_kable(table_kable, "emm_table.pdf")

# looking at intervention varibaility week to week

weekly_walks %>%
  filter(type == "Intervention") %>%
  group_by(week) %>%
  summarise(
    n = n(),
    mean = mean(weekly_planned_walks, na.rm = TRUE),
    sd = sd(weekly_planned_walks, na.rm = TRUE),
    median = median(weekly_planned_walks, na.rm = TRUE),
    IQR = IQR(weekly_planned_walks, na.rm = TRUE),
    min = min(weekly_planned_walks, na.rm = TRUE),
    max = max(weekly_planned_walks, na.rm = TRUE)
  )
```

### Proportion of walks that were planned

```{r}
# proportion of walks that were planned
walk_proportions <- all_walks %>%
  filter(type == "Intervention",
        walk_type != "No walk") %>%
  filter(week <=12) %>%
  count(walk_type) %>%
  mutate(
    proportion = n / sum(n),
    percentage = proportion * 100
  )

print(walk_proportions)
```

### Model 1: Plot raw data

```{r}
weekly_walks %>%
  group_by(type, week) %>%
  summarise(
    mean_walks = mean(weekly_walks, na.rm = TRUE),
    se_walks = sd(weekly_walks, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = week, y = mean_walks, color = type, group = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_walks - se_walks,
                    ymax = mean_walks + se_walks),
                width = 0.3) +
  # Add horizontal dashed line at Y=5
  geom_hline(yintercept = 5, linetype = "dashed", color = "black", linewidth = 0.8) +
  # Add significance stars above x=1, x=6, and x=12
  annotate("text", x = 1, y = Inf, label = "*",
           size = 6, vjust = 1.5, color = "black") +
  annotate("text", x = 6, y = Inf, label = "*",
           size = 6, vjust = 1.5, color = "black") +
  annotate("text", x = 12, y = Inf, label = "*",
           size = 6, vjust = 1.5, color = "black") +
  scale_color_manual(values = c("Control" = "#FFBE6A",
                                  "Intervention" = "#40B0A6")) +
  scale_y_continuous(breaks = seq(0, 7, by = 1)) +
  labs(
    title = "Weekly Total Walks Over Time",
    x = "Week",
    y = "Number of Walks (Mean ± SE)",
    color = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_x_continuous(breaks = 1:12)
```

### Model 2: Planned walks (Intervention) vs Total walks (Control) - Adjusted Model

```{r}
# Create outcome variable
df_long_model2 <- weekly_walks %>%
  mutate(walks_outcome = ifelse(type == "Intervention",
                                 weekly_planned_walks,    # For Intervention
                                 weekly_walks))           # For Control

# Adjusted Model
glmodel2_adj <- glmer(walks_outcome ~ type * week + age2 + sex_cat +
                      years_of_ed + days_since_injury_scaled + (1 + week | id),
                      control = glmerControl(optimizer = "bobyqa"),
                      data = df_long_model2,
                      family = poisson(link = "log"))
summary(glmodel2_adj)

# Extract IRRs and CIs
tidy(glmodel2_adj, effects = "fixed") %>%
  mutate(
    IRR = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error)
  )
```

### Model 2: Estimated marginal means at weeks 1, 6, 12

```{r}
# Estimated marginal means from adjusted model
emm2_adj <- emmeans(glmodel2_adj, ~ type | week, at = list(week = c(1, 6, 12)), type = "response")
print(emm2_adj)

# Pairwise comparisons: Intervention vs Control at each week
pairs(emm2_adj, by = "week")

# Extract EMMs
emm_table <- emm2_adj %>%
  as.data.frame() %>%
  select(week, type, rate) %>%
  pivot_wider(names_from = type, values_from = rate) %>%
  mutate(
    Ratio = Intervention / Control,
    Ratio_Label = paste0(round(Ratio, 1), "× more")
  )

print(emm_table)

emm_table_with_p <- emm_table %>%
  mutate(`P-value` = c("<0.001", "<0.001", "<0.001"))


# Create kable table
table_kable <- emm_table_with_p %>%
  kbl(caption = "Estimated Marginal Means of Planned (intervention) vs total Walks",
      booktabs = TRUE,
      align = "c",
      digits = 2) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)

table_kable
```

### Model 2: Plot

```{r}
df_long_model2 %>%
  group_by(type, week) %>%
  summarise(
    mean_walks = mean(walks_outcome, na.rm = TRUE),
    se_walks = sd(walks_outcome, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = week, y = mean_walks, color = type, group = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_walks - se_walks,
                    ymax = mean_walks + se_walks),
                width = 0.3) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "black", linewidth = 0.8) +
 scale_color_manual(values = c("Control" = "#FFBE6A",
                                  "Intervention" = "#40B0A6")) +
  scale_y_continuous(breaks = seq(0, 6, by = 1)) +
  labs(
    title = "Planned Walks (Intervention) vs Total Walks (Control) Over Time",
    x = "Week",
    y = "Number of Walks (Mean ± SE)",
    color = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_x_continuous(breaks = 1:12)
```

### Model 2: Visualization with individual trajectories

```{r}
df_long_model2 %>%
  ggplot(aes(x = week)) +
  # Individual participant trajectories (transparent)
  geom_line(aes(y = walks_outcome, group = id, color = type),
            alpha = 0.25, linewidth = 0.5, na.rm = TRUE) +
  geom_point(aes(y = walks_outcome, group = id, color = type),
             alpha = 0.25, size = 0.5, na.rm = TRUE) +
  # Mean trajectories (bold)
  stat_summary(aes(y = walks_outcome, group = type, color = type),
               fun = mean, geom = "line", linewidth = 1.5, na.rm = TRUE) +
  stat_summary(aes(y = walks_outcome, group = type, color = type),
               fun = mean, geom = "point", size = 3, na.rm = TRUE) +
  # Reference line at 5 walks
  geom_hline(yintercept = 5, linetype = "dashed", color = "black", linewidth = 0.8) +
  # Colors
  scale_color_manual(
    values = c("Control" = "#40B0A6", "Intervention" = "#D55E00"),
    labels = c("Control (Total Walks)", "Intervention (Planned Walks)")
  ) +
  # Axes
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(breaks = seq(0, 10, by = 1)) +
  # Labels
  labs(
    title = "Planned Walks (Intervention) vs Total Walks (Control) Over Time",
    subtitle = "Individual trajectories (transparent) and group means (bold)",
    x = "Week",
    y = "Number of Walks",
    color = "Group"
  ) +
  # Theme
  theme_minimal(base_size = 14, base_family = "Times New Roman") +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.key.size = unit(0.5, "cm"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.margin = margin(t = 15, r = 15, b = 15, l = 10)
  )
```

### Model 1 and 2 combined plot for model output - used in paper

```{r}
# Get EMMs from both models
emm_total <- emmeans(glmodel1_adj, ~ type | week,
                     at = list(week = 1:12),
                     type = "response") %>%
  as.data.frame() %>%
  mutate(walk_type = "Total")

emm_planned <- emmeans(glmodel2_adj, ~ type | week,
                       at = list(week = 1:12),
                       type = "response") %>%
  as.data.frame() %>%
  mutate(walk_type = "Planned")

# Combine but filter intervention planned
plot_data <- bind_rows(
  emm_total %>% filter(type == "Control"),  # Control total only
  emm_total %>% filter(type == "Intervention"),  # Intervention total
  emm_planned %>% filter(type == "Intervention")  # Intervention planned
) %>%
  mutate(
    group_label = case_when(
      type == "Control" ~ "Control (Total)",
      type == "Intervention" & walk_type == "Total" ~ "Intervention (Total)",
      type == "Intervention" & walk_type == "Planned" ~ "Intervention (Planned)"
    )
  )

# Create significance markers
sig_data <- data.frame(
  week = c(1, 6, 12),
  rate = c(
    max(plot_data$asymp.UCL[plot_data$week == 1]) + 0.5,
    max(plot_data$asymp.UCL[plot_data$week == 6]) + 0.5,
    max(plot_data$asymp.UCL[plot_data$week == 12]) + 0.5
  ),
  label = c("***", "***", "***")
)


# Plot
ggplot(plot_data, aes(x = week, y = rate, color = group_label, group = group_label)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, fill = group_label),
              alpha = 0.2, color = NA) +
 geom_text(data = sig_data, aes(x = week, y = rate, label = label),
            color = "black", size = 6, inherit.aes = FALSE) +
  labs(
    x = "Week",
    y = "Estimated Weekly Walks",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(breaks = 0:9) +
  scale_color_manual(values = c(
    "Control (Total)" = "#A8DCD1",
    "Intervention (Total)" = "#FFBE6A",
    "Intervention (Planned)" = "#D55E00"
  )) +
  scale_fill_manual(values = c(
    "Control (Total)" = "#A8DCD1",
    "Intervention (Total)" = "#FFBE6A",
    "Intervention (Planned)" = "#D55E00"
  ))
```

### Model 2: Bar plot with individual data points for each week

```{r}
# Bar plot with individual data points for each week planned vs total
df_long_model2 %>%
  ggplot(aes(x = factor(week), y = walks_outcome, fill = type)) +
  # Bar for means
  stat_summary(fun = mean, geom = "bar", alpha = 0.7,
               position = position_dodge(width = 0.9), color = "black") +
  # Individual data points
  geom_point(aes(color = type),
             position = position_jitterdodge(jitter.width = 0.2,
                                             dodge.width = 0.9),
             alpha = 0.6, size = 2) +
  # Error bars (SE)
  stat_summary(fun.data = mean_se, geom = "errorbar",
               position = position_dodge(width = 0.9),
               width = 0.2, linewidth = 0.8) +
  # Reference line at 5 walks
  geom_hline(yintercept = 5, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  # Colors
  scale_fill_manual(
    values = c("Control" = "#FFBE6A", "Intervention" = "#40B0A6"),
    labels = c("Control (Total Walks)", "Intervention (Planned Walks)")
  ) +
  scale_color_manual(
    values = c("Control" = "#FFBE6AC0", "Intervention" = "#40B0A640"),
    labels = c("Control (Total Walks)", "Intervention (Planned Walks)")
  ) +
  # Axes
  scale_y_continuous(breaks = seq(0, 10, by = 1)) +
  labs(
    title = "Planned (Intervention) vs Total Walks (Control)",
    x = "Week",
    y = "Number of Walks",
    fill = "Group",
    color = "Group"
  ) +
  # Theme
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom",
    panel.grid.major.x = element_blank()
  )
```

### Combined plot for both planned and unplanned - used in paper

```{r}
# --- Summarise means ---

df_int_summary <- weekly_walks %>%
  filter(type == "Intervention") %>%
  group_by(week) %>%
  summarise(
    planned_mean = mean(weekly_planned_walks, na.rm = TRUE),
    unplanned_mean = mean(weekly_unplanned_walks, na.rm = TRUE),
    .groups = "drop"
  )

df_ctrl_summary <- weekly_walks %>%
  filter(type == "Control") %>%
  group_by(week) %>%
  summarise(
    total_mean = mean(weekly_walks, na.rm = TRUE),
    .groups = "drop"
  )

# --- Build bar segments manually for proper stacking ---
bar_width <- 0.4

df_rects <- bind_rows(
  # Control bar
  df_ctrl_summary %>%
    mutate(
      xmin = as.numeric(factor(week)) - 0.225 - bar_width/2,
      xmax = as.numeric(factor(week)) - 0.225 + bar_width/2,
      ymin = 0,
      ymax = total_mean,
      fill_group = "Control (Total Walks)"
    ),
  # Intervention: planned (bottom)
  df_int_summary %>%
    mutate(
      xmin = as.numeric(factor(week)) + 0.225 - bar_width/2,
      xmax = as.numeric(factor(week)) + 0.225 + bar_width/2,
      ymin = 0,
      ymax = planned_mean,
      fill_group = "Intervention (Planned)"
    ),
  # Intervention: unplanned (top, stacked on planned)
  df_int_summary %>%
    mutate(
      xmin = as.numeric(factor(week)) + 0.225 - bar_width/2,
      xmax = as.numeric(factor(week)) + 0.225 + bar_width/2,
      ymin = planned_mean,
      ymax = planned_mean + unplanned_mean,
      fill_group = "Intervention (Unplanned)"
    )
)

df_rects$fill_group <- factor(df_rects$fill_group,
                              levels = c("Control (Total Walks)",
                                         "Intervention (Planned)",
                                         "Intervention (Unplanned)"))

# --- Individual dots ---

df_dots <- weekly_walks %>%
  mutate(
    bar_group = type,
    total = if_else(type == "Intervention",
                    weekly_planned_walks + weekly_unplanned_walks,
                    weekly_walks)
  )
df_dots$bar_group <- factor(df_dots$bar_group,
                            levels = c("Control", "Intervention"))

# --- Plot ---

ggplot() +
  geom_rect(data = df_rects,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
                fill = fill_group),
            color = "black", alpha = 0.85) +
  geom_point(data = df_dots,
             aes(x = factor(week), y = total,
                 colour = bar_group, fill = bar_group),
             position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.9),
             alpha = 0.5, size = 2, show.legend = FALSE) +
  geom_hline(yintercept = 5, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  scale_fill_manual(
    name = "",
    values = c(
      "Control (Total Walks)"    = "#A8DCD1",
      "Intervention (Planned)"   = "#D55E00",
      "Intervention (Unplanned)" = "#FFBE6A",
      "Control"                  = "#A8DCD1",
      "Intervention"             = "#FFBE6A"
    ),
    breaks = c("Control (Total Walks)",
               "Intervention (Planned)",
               "Intervention (Unplanned)")
  ) +
  scale_colour_manual(
    values = c("Control" = "#A8DCD1C0", "Intervention" = "#EA8E35"),
    guide = "none"
  ) +
  scale_x_discrete() +
  scale_y_continuous(breaks = seq(0, 20, by = 2)) +
  labs(
    title = "Weekly Walks",
    subtitle = "Bar height = mean total walks; dots = individual participant totals",
    x = "Week",
    y = "Number of Walks"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom",
    panel.grid.major.x = element_blank()
  )
```

### Model 2: Bar plot with individual data points for timepoint 1,6,and 12

```{r}
# Filter to weeks 1, 6, and 12
df_subset <- df_long_model2 %>%
  filter(week %in% c(1, 6, 12))

df_subset %>%
  ggplot(aes(x = factor(week), y = walks_outcome, fill = type)) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.7,
               position = position_dodge(width = 0.9),
               color = "black", linewidth = 0.5) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               position = position_dodge(width = 0.9),
               width = 0.25, linewidth = 0.8) +
  geom_point(aes(color = type),
             position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.9),
             alpha = 0.6, size = 2.5) +
  geom_hline(yintercept = 5, linetype = "dashed",
             color = "gray40", linewidth = 0.8) +
  scale_fill_manual(
    values = c("Control" = "#FFBE6A", "Intervention" = "#40B0A6"),
    labels = c("Control (Total)", "Intervention (Planned)")
  ) +
  scale_color_manual(
    values = c("Control" = "#1e40af", "Intervention" = "#991b1b"),
    labels = c("Control (Total)", "Intervention (Planned)")
  ) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), limits = c(0, 10)) +
  scale_x_discrete(labels = c("Week 1", "Week 6", "Week 12")) +
  labs(
    title = "Walking Activity: Control vs Intervention",
    x = "Time Point",
    y = "Number of Walks",
    fill = "Group",
    color = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Model 3: Total Duration of Walks - Adjusted Model

```{r}
# remove outlier
weekly_walks_out <- weekly_walks %>%
  filter(total_duration_walks > quantile(total_duration_walks, 0.25) - 1.5 * IQR(total_duration_walks) &
         total_duration_walks < quantile(total_duration_walks, 0.75) + 1.5 * IQR(total_duration_walks))

# Adjusted Model (race excluded due to limited sample diversity)
model3_adj <- lmer(total_duration_walks ~ type * week + age2 + sex_cat +
                   years_of_ed + days_since_injury_scaled + (1 | id),
                   data = weekly_walks_out,
                   REML = TRUE)

summary(model3_adj)

# Note: Random intercept only (1 | id) used due to convergence issues with random slope specification (1 + week | id)

# Extract coefficients with CIs
tidy(model3_adj, effects = "fixed", conf.int = TRUE)
```

### Model 3: Marginal means

```{r}
# Estimated marginal means from adjusted model
emm3_adj <- emmeans(model3_adj, ~ type | week, at = list(week = c(1, 6, 12)), type = "response")
print(emm3_adj)

# Pairwise comparisons: Intervention vs Control at each week
pairs(emm3_adj, by = "week")

# Extract EMMs
emm_table3 <- emm3_adj %>%
  as.data.frame() %>%
  select(week, type, emmean) %>%
  pivot_wider(names_from = type, values_from = emmean) %>%
  mutate(
    Difference = Intervention / Control,
    Diff_Label = paste0(round(Difference, 1), "× more")
  )

print(emm_table3)

emm_table3_with_p <- emm_table3 %>%
  mutate(`P-value` = c("<0.001", "<0.001", "<0.001"))

# Create kable table
table_kable <- emm_table3_with_p %>%
  kbl(caption = "Estimated Marginal Means of Weekly Walk Duration",
      booktabs = TRUE,
      align = "c",
      digits = 2) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)

table_kable
```

### Model 3: Visualization

```{r}
weekly_walks_out %>%
  group_by(type, week) %>%
  summarise(
    mean_duration = mean(total_duration_walks, na.rm = TRUE),
    se_duration = sd(total_duration_walks, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = week, y = mean_duration, color = type, group = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_duration - se_duration,
                    ymax = mean_duration + se_duration),
                width = 0.3) +
  scale_color_manual(values = c("Control" = "#40B0A6",
                                 "Intervention" = "#D55E00")) +
  labs(
    title = "Walking Duration Over Time",
    x = "Week",
    y = "Duration (minutes, Mean ± SE)",
    color = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_x_continuous(breaks = 1:12)

# violin plot

weekly_walks_out %>%
  ggplot(aes(x = as.factor(week), y = total_duration_walks, fill = type)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", size = 2,
               color = "black", position = position_dodge(0.9)) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.2, color = "black",
               position = position_dodge(0.9)) +
  scale_fill_manual(values = c("Control" = "#40B0A6",
                                "Intervention" = "#D55E00")) +
  labs(
    title = "Walking Duration Over Time",
    x = "Week",
    y = "Duration (minutes)",
    fill = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

# ridge plot
library(ggridges)

weekly_walks_out %>%
  ggplot(aes(x = total_duration_walks, y = as.factor(week), fill = type, weight = weekly_walks)) +
  geom_density_ridges(alpha = 0.7, scale = 0.9) +
  scale_fill_manual(values = c("Control" = "#40B0A6",
                                "Intervention" = "#D55E00")) +
scale_y_discrete(limits = rev) +
  labs(
    title = "Walking Duration Over Time",
    x = "Duration (minutes)",
    y = "Week",
    fill = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

# ribbon plot — used in paper
p_means <- df_means %>%
  ggplot(aes(week, duration_mean, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = duration_mean - duration_se,
                  ymax = duration_mean + duration_se), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Control" = "#A8DCD1", "Intervention" = "#FFBE6A")) +
  scale_fill_manual(values = c("Control" = "#A8DCD1", "Intervention" = "#FFBE6A")) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Week", y = "Duration (mins)", colour = NULL, fill = NULL) +
  annotate("text", x = c(1, 6, 12), y = Inf, label = "***", vjust = 1.5, size = 5) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank()
  )
p_means
```

**MODEL 3 SUMMARY: Total Walking Duration**

**Main Finding:**

Intervention group walks 331 more minutes per week than Control (95% CI: -11 to 673 minutes), approximately 5-6 extra hours per week. This exceeds WHO's recommended 150 minutes of moderate activity by more than double. The wide confidence interval reflects substantial individual variability in walking duration.

### Model 4: Steps in Walks - Adjusted Model

```{r}
# remove outlier (applied on top of Model 3 outlier-filtered dataset)
weekly_walks_out <- weekly_walks_out %>%
  filter(total_steps_in_walks > quantile(total_steps_in_walks, 0.25) - 1.5 * IQR(total_steps_in_walks) &
         total_steps_in_walks < quantile(total_steps_in_walks, 0.75) + 1.5 * IQR(total_steps_in_walks))

# Adjusted model
model4_adj <- lmer(total_steps_in_walks ~ type * week + age2 + sex_cat +
                   years_of_ed + days_since_injury_scaled + (1 | id),
                   data = weekly_walks_out,
                   REML = TRUE)
summary(model4_adj)

# Extract coefficients with CIs
library(broom.mixed)
tidy(model4_adj, effects = "fixed", conf.int = TRUE)
```

### Model 4: Marginal means

```{r}
# Estimated marginal means from adjusted model
emm4_adj <- emmeans(model4_adj, ~ type | week, at = list(week = c(1, 6, 12)), type = "response")
print(emm4_adj)

# Pairwise comparisons: Intervention vs Control at each week
pairs(emm4_adj, by = "week")

# Extract EMMs
emm_table4 <- emm4_adj %>%
  as.data.frame() %>%
  select(week, type, emmean) %>%
  pivot_wider(names_from = type, values_from = emmean) %>%
  mutate(
    Difference = Intervention / Control,
    Diff_Label = paste0(round(Difference, 1), "× more")
  )

print(emm_table4)

emm_table4_with_p <- emm_table4 %>%
  mutate(`P-value` = c("<0.001", "<0.001", "<0.001"))

# Create kable table
table_kable <- emm_table4_with_p %>%
  kbl(caption = "Estimated Marginal Means of Weekly step count during walks",
      booktabs = TRUE,
      align = "c",
      digits = 2) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)

table_kable
```

### Model 4: Visualization

```{r}
weekly_walks_out %>%
  group_by(type, week) %>%
  summarise(
    mean_steps = mean(total_steps_in_walks, na.rm = TRUE),
    se_steps = sd(total_steps_in_walks, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = week, y = mean_steps, color = type, group = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_steps - se_steps,
                    ymax = mean_steps + se_steps),
                width = 0.3) +
  scale_color_manual(values = c("Control" = "#40B0A6",
                                 "Intervention" = "#D55E00")) +
  labs(
    title = "Steps in Walks Over Time",
    x = "Week",
    y = "Steps (Mean ± SE)",
    color = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_x_continuous(breaks = 1:12)

# ribbon plot — used in paper
p_means_step <- df_means %>%
  ggplot(aes(week, total_steps_in_walks_mean, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = total_steps_in_walks_mean - total_steps_in_walks_se,
                  ymax = total_steps_in_walks_mean + total_steps_in_walks_se), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Control" = "#A8DCD1", "Intervention" = "#FFBE6A")) +
  scale_fill_manual(values = c("Control" = "#A8DCD1", "Intervention" = "#FFBE6A")) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Week", y = "Step Count", colour = NULL, fill = NULL) +
  annotate("text", x = c(1, 6, 12), y = Inf, label = "***", vjust = 1.5, size = 5) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank()
  )
p_means_step
```

**MODEL 4 SUMMARY: Steps in Walks - Weeks 1-12**

**Main Finding:**

Intervention group accumulates 8,962 more steps per week in walks than Control (95% CI: -4,414 to 22,337 steps). The wide confidence interval reflects substantial individual variability in walking patterns.

```{r}
# trying a single plot for duraiton and steps

ggplot(weekly_walks_out, aes(x = week, y = total_duration_walks, size = total_steps_in_walks, color = type)) +
  geom_point(alpha = 0.6) +
  labs(x = "Week", y = "Total Duration (min)", size = "Total Steps",
       title = "Weekly Activity: Duration and Steps")
```