---
title: "Roybal - Feasibility & Compliance"
format:
  html:
    toc: true
    embed-resources: true
editor: visual
editor_options:
  chunk_output_group: inline
  author: "Timothy P Morris"
---

# Roybal Final Data Analysis — Feasibility & Compliance

**Purpose:** This document covers feasibility of intervention delivery and study procedures, and participant compliance outcomes from the Roybal pilot RCT\~

------------------------------------------------------------------------

### Load libraries

```{r}
pkgs <- c(
  "readr", "readxl", "skimr", "tidyverse", "lme4", "lmerTest",
  "tidymodels", "performance", "emmeans", "pwr", "longpower", "simr",
  "ggplot2", "consort", "kableExtra", "broom.mixed", "effectsize",
  "car", "gridExtra", "ggstatsplot", "ggeffects", "tableone",
  "ggpubr", "flextable", "officer"
)

install.packages(setdiff(pkgs, rownames(installed.packages())))

invisible(lapply(pkgs, library, character.only = TRUE))
```

### Load data

```{r}
load("roybal_base.RData")

# Fix weather id capitalisation and join weekly summary cols to weekly_walks
weather_clean <- weather_raw %>%
  mutate(id = gsub("^roybal", "Roybal", id))

weekly_walks <- weekly_walks %>%
  left_join(
    weather_clean %>% select(id, week, avg_temp_f, avg_precip_in, pct_precip_days, total_precip_in),
    by = c("id", "week")
  )
```

## Feasibility

### Consort Diagram

```{r}
library(consort)

consort <- add_box(txt = "Inquired (n = 341)")

consort <- add_side_box(consort,
                  txt = "Excluded (n = 223)\n• Not meeting criteria (n = 62)\n• Lost to follow up (n = 159)\n• Declined to participate (n = 2)")

consort <- add_box(consort, txt = "Pre-screened (n = 118)")

consort <- add_side_box(consort,
                  txt = "Excluded (n = 62)\n• Not meeting criteria (n = 32)\n• Lost to follow up (n = 27)\n• Declined to participate (n = 3)")

consort <- add_box(consort, txt = "Randomized (n = 56)")

consort <- add_split(consort,
                     txt = c("Intervention (n = 28)",
                            "Control (n = 28)"))

consort <- add_side_box(consort,
                  txt = c("Dropout (n = 5)",
                          "Dropout (n = 3)"))

consort <- add_box(consort,
                   txt = c("Intervention Complete (n = 23)",
                          "Control Complete (n = 25)"))

# Plot
plot(consort)
```

### Table One

```{r}
library(readxl)
library(dplyr)
library(tableone)
library(knitr)
library(kableExtra)

# Load datasets
completed <- read_excel("final_datasets/roybal_masterdoc.xlsx", sheet = 1)
dropped <- read_excel("final_datasets/roybal_masterdoc.xlsx", sheet = 2)

# Keep only these dropout IDs
include_ids <- c("Roybal_08", "Roybal_09", "Roybal_10", "Roybal_20",
                 "Roybal_22", "Roybal_23", "Roybal_37", "Roybal_42", "Roybal_55")
dropped_filtered <- dropped %>% filter(id %in% include_ids)

# Rename variables
completed_clean <- completed %>%
  rename(
    `Age (years)` = age2,
    `Biological Sex` = sex_cat,
    `Gender Identity` = gender_cat,
    Race = race_cat,
    Ethnicity = ethnicity_cat,
    `Time Since Injury (days)` = days_since_injury,
    `Loss of Consciousness` = loc,
    `Memory Gap` = memory_gap
  )

dropped_clean <- dropped_filtered %>%
  rename(
    `Age (years)` = age2,
    `Biological Sex` = sex_cat,
    `Gender Identity` = gender_cat,
    Race = race_cat,
    Ethnicity = ethnicity_cat,
    `Time Since Injury (days)` = days_since_injury,
    `Loss of Consciousness` = loc,
    `Memory Gap` = memory_gap
  )

# Combine datasets
df_table1 <- bind_rows(
  completed_clean %>% mutate(group_status = group),
  dropped_clean %>% mutate(group_status = "Dropout")
)

# Set factor order
df_table1$group_status <- factor(df_table1$group_status,
                                  levels = c("Intervention", "Control", "Dropout"))

# Define variables
vars <- c("Age (years)",
          "Biological Sex",
          "Gender Identity",
          "Race",
          "Ethnicity",
          "Time Since Injury (days)",
          "Loss of Consciousness",
          "Memory Gap")

catVars <- c("Biological Sex", "Gender Identity", "Race", "Ethnicity",
             "Loss of Consciousness", "Memory Gap")

# Create Table 1
table1 <- CreateTableOne(
  vars = vars,
  strata = "group_status",
  data = df_table1,
  factorVars = catVars,
  test = TRUE,
  addOverall = TRUE
)

# Print with formatting
table1_print <- print(table1,
                      nonnormal = NULL,
                      exact = NULL,
                      quote = FALSE,
                      noSpaces = TRUE,
                      printToggle = FALSE,
                      showAllLevels = TRUE,
                      contDigits = 1,
                      catDigits = 1,
                      pDigits = 3,
                      minMax = TRUE)

# Make final table
kable(table1_print,
      caption = "Table 1: Baseline Characteristics by Group",
      align = "lccccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "15em")

  # Convert to data frame first
table1_df <- as.data.frame(table1_print)


# Convert to flextable
final_table <- flextable(table1_df) %>%
  set_caption("Table 1: Baseline Characteristics by Group") %>%
  bold(j = 1, part = "body") %>%  # Bold first column
  bold(part = "header") %>%  # Bold headers
  fontsize(size = 12, part = "all") %>%
  autofit()

# Save as Word
save_as_docx(final_table, path = "table1.docx")
```

### Dropout and Retention Rate

```{r}
dropped_real <- dropped %>%
  filter(id %in% c("Roybal_08", "Roybal_09", "Roybal_10", "Roybal_20",
                   "Roybal_22", "Roybal_23", "Roybal_37", "Roybal_42", "Roybal_55"))

n_enrolled <- nrow(completed) + nrow(dropped_real)
n_completed <- nrow(completed)
n_dropped <- nrow(dropped_real)

data.frame(
  Category = c("Total Enrolled", "Completed", "Dropped Out"),
  N = c(n_enrolled, n_completed, n_dropped),
  Percentage = paste0(round(c(100, n_completed/n_enrolled*100, n_dropped/n_enrolled*100), 1), "%")
) %>%
  kable(caption = "Enrollment and Retention") %>%
  kable_styling()
```

### Dropout Rate Visualization

```{r}
dropped_real <- dropped %>%
  filter(id %in% c("Roybal_08", "Roybal_09", "Roybal_10", "Roybal_20",
                   "Roybal_22", "Roybal_23", "Roybal_37", "Roybal_42", "Roybal_55"))

n_enrolled <- nrow(completed) + nrow(dropped_real)
n_completed <- nrow(completed)
n_dropped <- nrow(dropped_real)

data.frame(
  Category = factor(c("Enrolled", "Completed", "Dropped Out"),
                   levels = c("Enrolled", "Completed", "Dropped Out")),
  N = c(n_enrolled, n_completed, n_dropped),
  Percentage = round(c(100, n_completed/n_enrolled*100, n_dropped/n_enrolled*100), 1)
) %>%
  ggplot(aes(x = Category, y = N, fill = Category)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = paste0(N, "\n(", Percentage, "%)")),
            vjust = -0.3, fontface = "bold", size = 5) +
  scale_fill_manual(values = c("Enrolled" = "#3b82f6",
                                "Completed" = "#16a34a",
                                "Dropped Out" = "#ef4444")) +
  labs(
    title = "Study Enrollment and Retention",
    x = "",
    y = "Number of Participants"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  ylim(0, n_enrolled + 5)
```

### Days between Baseline and Fitbit Setup/Call 0

```{r}
completed %>%
  group_by(group) %>%
  summarise(N = n(),
            `Mean (SD)` = paste0(round(mean(days_between_baseline_call0, na.rm=T), 1),
                                " (", round(sd(days_between_baseline_call0, na.rm=T), 1), ")")) %>%
  bind_rows(completed %>%
              summarise(group = "Overall", N = n(),
                       `Mean (SD)` = paste0(round(mean(days_between_baseline_call0, na.rm=T), 1),
                                           " (", round(sd(days_between_baseline_call0, na.rm=T), 1), ")"))) %>%
  kable(caption = "Days Between Baseline and Fitbit Setup") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Days between Call 11 and Endpoint

```{r}
completed %>%
  group_by(group) %>%
  summarise(N = n(),
            `Mean (SD)` = paste0(round(mean(days_between_call11_endpoint, na.rm=T), 1),
                                " (", round(sd(days_between_call11_endpoint, na.rm=T), 1), ")")) %>%
  bind_rows(completed %>%
              summarise(group = "Overall", N = n(),
                       `Mean (SD)` = paste0(round(mean(days_between_call11_endpoint, na.rm=T), 1),
                                           " (", round(sd(days_between_call11_endpoint, na.rm=T), 1), ")"))) %>%
  kable(caption = "Days Between Call 11 and Endpoint") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Compliance

### Phone Calls Made

```{r}
completed %>%
  group_by(group) %>%
  summarise(
    N = n(),
    `Completed Mean (SD)` = paste0(round(mean(total_completed_calls, na.rm=T), 1),
                                   " (", round(sd(total_completed_calls, na.rm=T), 1), ")"),
    `Completion Rate (%)` = paste0(round(mean(total_completed_calls/total_calls * 100, na.rm=T), 1), "%"),
    `Missed Mean (SD)` = paste0(round(mean(total_missed_calls, na.rm=T), 1),
                                " (", round(sd(total_missed_calls, na.rm=T), 1), ")")
  ) %>%
  bind_rows(completed %>%
              summarise(
                group = "Overall",
                N = n(),
                `Completed Mean (SD)` = paste0(round(mean(total_completed_calls, na.rm=T), 1),
                                               " (", round(sd(total_completed_calls, na.rm=T), 1), ")"),
                `Completion Rate (%)` = paste0(round(mean(total_completed_calls/total_calls * 100, na.rm=T), 1), "%"),
                `Missed Mean (SD)` = paste0(round(mean(total_missed_calls, na.rm=T), 1),
                                            " (", round(sd(total_missed_calls, na.rm=T), 1), ")")
              )) %>%
  kable(caption = "Phone Call Compliance (Total Scheduled = 11 calls)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Completion Rate Pie Chart

```{r}
library(ggpubr)

# Create data for pie charts
pie_data <- completed %>%
  group_by(group) %>%
  summarise(
    Completed = sum(total_completed_calls, na.rm = TRUE),
    Missed = sum(total_missed_calls, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Completed, Missed), names_to = "Status", values_to = "Count") %>%
  group_by(group) %>%
  mutate(Percentage = round((Count / sum(Count)) * 100, 1))

# Control pie
p1 <- pie_data %>%
  filter(group == "Control") %>%
  ggpie("Count", label = paste0(pie_data$Count[pie_data$group=="Control"],
                                 " (", pie_data$Percentage[pie_data$group=="Control"], "%)"),
        fill = "Status", color = "white",
        palette = c("#16a34a", "#D55E00"),
        title = "Control")

# Intervention pie
p2 <- pie_data %>%
  filter(group == "Intervention") %>%
  ggpie("Count", label = paste0(pie_data$Count[pie_data$group=="Intervention"],
                                 " (", pie_data$Percentage[pie_data$group=="Intervention"], "%)"),
        fill = "Status", color = "white",
        palette = c("#16a34a", "#D55E00"),
        title = "Intervention")

# Combine
ggarrange(p1, p2, ncol = 2, common.legend = TRUE, legend = "bottom")
```

### Average Phone Call Length by Group

```{r}
# re-do with IQR
completed %>%
  mutate(
    avg_call_minutes = sapply(average_call_length, function(x) {
      parts <- as.numeric(strsplit(as.character(x), ":")[[1]])
      if(length(parts) == 3) parts[1]*60 + parts[2] + parts[3]/60
      else if(length(parts) == 2) parts[1] + parts[2]/60
      else NA
    })
  ) %>%
  group_by(group) %>%
  summarise(
    N = n(),
    `Mean (SD)` = paste0(
      round(mean(avg_call_minutes, na.rm=T), 1),
      " (", round(sd(avg_call_minutes, na.rm=T), 1), ")"
    ),
    `Median (IQR)` = paste0(
      round(median(avg_call_minutes, na.rm=T), 1),
      " (", round(quantile(avg_call_minutes, 0.25, na.rm=T), 1),
      "-", round(quantile(avg_call_minutes, 0.75, na.rm=T), 1), ")"
    )
  ) %>%
  bind_rows(
    completed %>%
      mutate(
        avg_call_minutes = sapply(average_call_length, function(x) {
          parts <- as.numeric(strsplit(as.character(x), ":")[[1]])
          if(length(parts) == 3) parts[1]*60 + parts[2] + parts[3]/60
          else if(length(parts) == 2) parts[1] + parts[2]/60
          else NA
        })
      ) %>%
      summarise(
        group = "Overall",
        N = n(),
        `Mean (SD)` = paste0(
          round(mean(avg_call_minutes, na.rm=T), 1),
          " (", round(sd(avg_call_minutes, na.rm=T), 1), ")"
        ),
        `Median (IQR)` = paste0(
          round(median(avg_call_minutes, na.rm=T), 1),
          " (", round(quantile(avg_call_minutes, 0.25, na.rm=T), 1),
          "-", round(quantile(avg_call_minutes, 0.75, na.rm=T), 1), ")"
        )
      )
  ) %>%
  kable(caption = "Average Phone Call Length") %>%
  kable_styling()
```

### Average Call Length by Group: GGstats plot

```{r}
library(ggstatsplot)

# Boxplot with stats: Average Call Length by Group
completed %>%
  mutate(
    avg_call_minutes = sapply(average_call_length, function(x) {
      parts <- as.numeric(strsplit(as.character(x), ":")[[1]])
      if(length(parts) == 3) parts[1]*60 + parts[2] + parts[3]/60
      else if(length(parts) == 2) parts[1] + parts[2]/60
      else NA
    })
  ) %>%
  ggbetweenstats(
    x = group,
    y = avg_call_minutes,
    title = "Average Call Length by Group",
    xlab = "Group",
    ylab = "Average Call Length (minutes)",
    type = "non-parametric",
    bf.message = FALSE,
    pairwise.display = "significant",
    package = "ggsci",
    palette = "default_jco"
  )
```

### Fitbit Wear Time Compliance

```{r}
completed %>%
  group_by(group) %>%
  summarise(
    N = n(),
    `Days with Data Mean (SD)` = paste0(
      round(mean(days_with_data, na.rm=T), 1),
      " (", round(sd(days_with_data, na.rm=T), 1), ")"
    ),
    `% of Expected Days` = paste0(
      round(mean(percent_days_with_data, na.rm=T), 1), "%"
    ),
    `Mean Daily Wear Time (hours)` = paste0(
      round(mean(mean_wear_time_hours, na.rm=T), 1),
      " (", round(sd(mean_wear_time_hours, na.rm=T), 1), ")"
    ),
    `Valid Days ≥10hr Mean% (SD)` = paste0(
      round(mean(percent_valid_min10hr, na.rm=T), 1),
      " (", round(sd(percent_valid_min10hr, na.rm=T), 1), ")"
    ),
    `Daytime Wear % Mean (SD)` = paste0(
      round(mean(mean_percent_wear_during_day, na.rm=T), 1),
      " (", round(sd(mean_percent_wear_during_day, na.rm=T), 1), ")"
    ),
    `Nighttime Wear % Mean (SD)` = paste0(
      round(mean(mean_percent_wear_during_night, na.rm=T), 1),
      " (", round(sd(mean_percent_wear_during_night, na.rm=T), 1), ")"
    )
  ) %>%
  bind_rows(completed %>%
              summarise(
                group = "Overall",
                N = n(),
                `Days with Data Mean (SD)` = paste0(
                  round(mean(days_with_data, na.rm=T), 1),
                  " (", round(sd(days_with_data, na.rm=T), 1), ")"
                ),
                `% of Expected Days` = paste0(
                  round(mean(percent_days_with_data, na.rm=T), 1), "%"
                ),
                `Mean Daily Wear Time (hours)` = paste0(
                  round(mean(mean_wear_time_hours, na.rm=T), 1),
                  " (", round(sd(mean_wear_time_hours, na.rm=T), 1), ")"
                ),
                `Valid Days ≥10hr Mean% (SD)` = paste0(
                  round(mean(percent_valid_min10hr, na.rm=T), 1),
                  " (", round(sd(percent_valid_min10hr, na.rm=T), 1), ")"
                ),
                `Daytime Wear % Mean (SD)` = paste0(
                  round(mean(mean_percent_wear_during_day, na.rm=T), 1),
                  " (", round(sd(mean_percent_wear_during_day, na.rm=T), 1), ")"
                ),
                `Nighttime Wear % Mean (SD)` = paste0(
                  round(mean(mean_percent_wear_during_night, na.rm=T), 1),
                  " (", round(sd(mean_percent_wear_during_night, na.rm=T), 1), ")"
                )
              )) %>%
  kable(caption = "Fitbit Wear Time Compliance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Daily Wear Time by Group: GGstats plot

```{r}
library(ggstatsplot)

ggbetweenstats(
  data = completed,
  x = group,
  y = mean_wear_time_hours,
  title = "Daily Fitbit Wear Time by Group",
  xlab = "Group",
  ylab = "Mean Daily Wear Time (hours)",
  type = "non-parametric",
  var.equal = FALSE,
  bf.message = FALSE,
  pairwise.display = "significant",
  package = "ggsci",
  palette = "default_jco"
)
```

### Check normality for all continuous compliance variables

```{r}
library(ggplot2)

# 1. Average Call Length
completed_calls <- completed %>%
  mutate(
    avg_call_minutes = sapply(average_call_length, function(x) {
      parts <- as.numeric(strsplit(as.character(x), ":")[[1]])
      if(length(parts) == 3) parts[1]*60 + parts[2] + parts[3]/60
      else if(length(parts) == 2) parts[1] + parts[2]/60
      else NA
    })
  )

# Histograms and Q-Q plots
par(mfrow = c(2, 3))

# Average call length
hist(completed_calls$avg_call_minutes[completed_calls$group == "Control"],
     main = "Control: Avg Call Length", xlab = "Minutes")
hist(completed_calls$avg_call_minutes[completed_calls$group == "Intervention"],
     main = "Intervention: Avg Call Length", xlab = "Minutes")

# Mean daily wear time
hist(completed$mean_wear_time_hours[completed$group == "Control"],
     main = "Control: Daily Wear Time", xlab = "Hours")
hist(completed$mean_wear_time_hours[completed$group == "Intervention"],
     main = "Intervention: Daily Wear Time", xlab = "Hours")

# Days with data
hist(completed$days_with_data[completed$group == "Control"],
     main = "Control: Days with Data", xlab = "Days")
hist(completed$days_with_data[completed$group == "Intervention"],
     main = "Intervention: Days with Data", xlab = "Days")

par(mfrow = c(1, 1))

# Shapiro-Wilk normality tests
cat("\n=== NORMALITY TESTS (Shapiro-Wilk) ===\n")
cat("p > 0.05 = normal distribution\n\n")

cat("Average Call Length:\n")
cat("  Control:", shapiro.test(completed_calls$avg_call_minutes[completed_calls$group == "Control"])$p.value, "\n")
cat("  Intervention:", shapiro.test(completed_calls$avg_call_minutes[completed_calls$group == "Intervention"])$p.value, "\n\n")

cat("Mean Daily Wear Time:\n")
cat("  Control:", shapiro.test(completed$mean_wear_time_hours[completed$group == "Control"])$p.value, "\n")
cat("  Intervention:", shapiro.test(completed$mean_wear_time_hours[completed$group == "Intervention"])$p.value, "\n\n")

cat("Days with Data:\n")
cat("  Control:", shapiro.test(completed$days_with_data[completed$group == "Control"])$p.value, "\n")
cat("  Intervention:", shapiro.test(completed$days_with_data[completed$group == "Intervention"])$p.value, "\n")
```

### Non-Parametric Tests (Mann-Whitney U)

```{r}
# 1. Average Call Length
completed_calls <- completed %>%
  mutate(
    avg_call_minutes = sapply(average_call_length, function(x) {
      parts <- as.numeric(strsplit(as.character(x), ":")[[1]])
      if(length(parts) == 3) parts[1]*60 + parts[2] + parts[3]/60
      else if(length(parts) == 2) parts[1] + parts[2]/60
      else NA
    })
  )

test1 <- wilcox.test(avg_call_minutes ~ group, data = completed_calls)
cat("Average Call Length: W =", test1$statistic, ", p =", round(test1$p.value, 3), "\n")

# 2. Mean Daily Wear Time
test2 <- wilcox.test(mean_wear_time_hours ~ group, data = completed)
cat("Mean Daily Wear Time: W =", test2$statistic, ", p =", round(test2$p.value, 3), "\n")

# 3. Days with Data
test3 <- wilcox.test(days_with_data ~ group, data = completed)
cat("Days with Data: W =", test3$statistic, ", p =", round(test3$p.value, 3), "\n")

# 4. Valid Days ≥10hr
test4 <- wilcox.test(percent_valid_min10hr ~ group, data = completed)
cat("Valid Days ≥10hr (%): W =", test4$statistic, ", p =", round(test4$p.value, 3), "\n")

# 5. Valid Days ≥80%
test5 <- wilcox.test(percent_valid_wear80 ~ group, data = completed)
cat("Valid Days ≥80% (%): W =", test5$statistic, ", p =", round(test5$p.value, 3), "\n")
```