---
title: "Modifiers"
format: html
author: "Timothy P Morris"
---

# Roybal Final Data Analysis — Modifiers

**Purpose:** This document covers modifier analyses of walking outcomes from the Roybal pilot RCT\~

## Load data

```{r}
load("roybal_base.RData")

# environment_vars — derived from completed (replaces broken original file path)
environment_vars <- completed %>%
  select(id, pc_park, urban_type, NatWalkInd)

# All six BREQ-3 regulation subscales, pre and post. All twelve are already present and
# fully populated (48/48) in `completed` straight from final_datasets/roybal_masterdoc.xlsx
# — build_data.R reads that sheet wholesale, so nothing needs adding upstream. Earlier
# versions of this chunk selected only intrinsic + amotivation, which is why the rest
# never reached the analysis dataframes.
breq3_subscales <- c("amotivation", "external", "introjected",
                     "identified", "integrated", "intrinsic")
breq3_vars <- c(paste0("breq3_", breq3_subscales, "_baseline"),
                paste0("breq3_", breq3_subscales, "_endpoint"))

base_vars <- completed %>%
  select(id, cog_func_tscore, gph_tscore, gmh_tscore, total_MET_min,
         all_of(breq3_vars),
         selfeff_total_score_baseli, age2, sex_cat, years_of_ed, days_since_injury)

mod_vars <- left_join(base_vars, environment_vars)

# Scale person-level vars
mod_vars <- mod_vars %>%
  mutate(across(c(NatWalkInd, pc_park, cog_func_tscore, gph_tscore, gmh_tscore,
                  all_of(breq3_vars),
                  selfeff_total_score_baseli, age2, years_of_ed, days_since_injury),
                ~ as.numeric(scale(.)),
                .names = "{.col}_z"))

# Process daily weather — take first walk record per day, filter to weeks 1-12, scale within-participant
dailyweather_filt <- weather_raw %>%
  arrange(participant_id, day, desc(has_walk_at_least_27min == "true")) %>%
  group_by(participant_id, day) %>%
  slice(1) %>%
  ungroup() %>%
  filter(week <= 12) %>%
  group_by(participant_id) %>%
  mutate(across(c(avg_temp_f, precip_in, cloud_cover_pct), ~ as.numeric(scale(.)))) %>%
  ungroup()

# Merge baseline vars to daily df
dailyweather_filt$participant_id <- tolower(dailyweather_filt$participant_id)
mod_vars$id <- tolower(mod_vars$id)

mods_final <- left_join(dailyweather_filt, mod_vars, by = c("participant_id" = "id")) %>%
  distinct()

mods_final <- mods_final %>%
  mutate(
    has_walk_at_least_27min = factor(has_walk_at_least_27min, levels = c("False", "True")),
    has_completed_walk      = factor(has_completed_walk,      levels = c("False", "True")),
    participant_id          = factor(participant_id)
  )

# create season var

  weekly_walks <- weekly_walks %>%
  mutate(
    midpoint = as.Date(official_start_date) + 
               as.numeric(as.Date(official_end_date) - as.Date(official_start_date)) / 2,
    mid_md = as.integer(format(midpoint, "%m%d")),
    season = case_when(
      mid_md >= 320  & mid_md < 621  ~ "Spring",
      mid_md >= 621  & mid_md < 923  ~ "Summer",
      mid_md >= 923  & mid_md < 1221 ~ "Fall",
      mid_md >= 1221 | mid_md < 320  ~ "Winter"
    )
  ) %>%
  select(-midpoint, -mid_md)


#extract one season value per participant and join to mods_final
season_lookup <- weekly_walks %>%
  distinct(id, season) %>%
  mutate(id = tolower(id))

mods_final <- left_join(mods_final, season_lookup, by = c("participant_id" = "id"))

# mods_final <- mods_final %>% select(-season.y)

# mods_final <- mods_final %>%
#    rename(season = `season.x`)

mods_final$season <- factor(mods_final$season, levels = c("Winter", "Spring", "Summer", "Fall"))

  
  
```

### models

```{r}
library(lme4)
library(broom.mixed)

# All candidate demographics (using scaled _z versions)
demographics <- c("age2_z", "sex_cat", "years_of_ed_z", "days_since_injury_z", "type", "month")

# Opportunity model: environmental vars
# avg_temp_f, precip_in, cloud_cover_pct already within-person scaled in mods_final
# NatWalkInd_z, pc_park_z are person-level scaled; urban_type is categorical
env_vars <- c("avg_temp_f", "precip_in", "cloud_cover_pct",
              "NatWalkInd_z", "pc_park_z", "urban_type", "season")

# Capability model: psychosocial vars
psy_vars <- c("gph_tscore", "gmh_tscore", "cog_func_tscore")

# Motivation model: motivational vars
mot_vars <- c("breq3_intrinsic_baseline_z", "breq3_amotivation_baseline_z",
              "selfeff_total_score_baseli_z")

fit_glmer <- function(predictors, label) {
  f <- paste("has_walk_at_least_27min ~",
             paste(predictors, collapse = " + "),
             "+ (1 | participant_id)")
  cat(sprintf("\n=== %s model ===\n", label))
  cat(sprintf("Formula: %s\n", f))
  tryCatch(
    glmer(as.formula(f), data = mods_final, family = binomial,
          control = glmerControl(optimizer = "bobyqa")),
    error = function(e) { cat("Model failed:", conditionMessage(e), "\n"); NULL }
  )
}

# --- Step 1: Demographics-only model ---
# Fit demographics alone to identify which covariates are independently
# associated with the outcome before adding domain-specific predictors.
mod_demographics <- fit_glmer(demographics, "Demographics only")
cat("\n\n========== Demographics-only Model Summary ==========\n")
print(summary(mod_demographics))

# Extract significant demographic terms (p < .05, excluding intercept).
# Strip dummy suffixes (e.g. "typeB" → "type") to recover the original
# variable name so the whole factor is retained or dropped together.
sig_demo_terms <- tidy(mod_demographics, effects = "fixed") %>%
  filter(term != "(Intercept)", p.value < 0.05) %>%
  pull(term)

sig_demographics <- demographics[sapply(demographics,
                                        function(v) any(startsWith(sig_demo_terms, v)))]

cat(sprintf("\nSignificant demographic(s) (p < .05): %s\n",
            paste(sig_demographics, collapse = ", ")))
cat("These will be carried forward into the domain models.\n")

# --- Step 2: Domain models with only significant demographics ---
mod_opportunity <- fit_glmer(c(env_vars, sig_demographics), "Opportunity")
mod_capability  <- fit_glmer(c(psy_vars, sig_demographics),  "Capability")
mod_motivation  <- fit_glmer(c(mot_vars, sig_demographics),  "Motivation")

# Summaries
for (nm in c("Opportunity", "Capability", "Motivation")) {
  m <- get(paste0("mod_", tolower(nm)))
  cat(sprintf("\n\n========== %s Model Summary ==========\n", nm))
  if (!is.null(m)) print(summary(m)) else cat("  Model unavailable.\n")
}

# Odds ratio tables
or_tables <- lapply(
  list(Opportunity = mod_opportunity,
       Capability  = mod_capability,
       Motivation  = mod_motivation),
  function(m) {
    if (is.null(m)) return(NULL)
    tidy(m, effects = "fixed", exponentiate = TRUE, conf.int = TRUE) %>%
      filter(term != "(Intercept)")
  }
)

cat("\n\n========== Opportunity OR Table ==========\n"); print(or_tables$Opportunity)
cat("\n========== Capability OR Table ==========\n");  print(or_tables$Capability)
cat("\n========== Motivation OR Table ==========\n");  print(or_tables$Motivation)
```

### Plots

```{r}
library(ggplot2)

# Combine OR tables from the three domain models into one dataframe.
# or_tables is a named list (Opportunity, Capability, Motivation) already
# extracted in the models chunk above with exponentiate = TRUE, conf.int = TRUE.
# Publication labels: maps tidy() term names to display labels.
# Factor dummies follow the pattern <varname><level> (e.g. "seasonSummer").
# If a term name differs in your data, update the left-hand side accordingly.
term_labels <- c(
  "typeIntervention"             = "Intervention",
  "pc_park_z"                    = "%Greenspace",
  "seasonSummer"                 = "Summer (v Winter)",
  "cloud_cover_pct"              = "%Cloud cover",
  "avg_temp_f"                   = "Daily temperature (°F)",
  "urban_typeUrban"              = "Urban v Rural",
  "precip_in"                    = "Daily precipitation (mm)",
  "NatWalkInd_z"                 = "Walkability Index",
  "seasonFall"                   = "Fall (v Winter)",
  "seasonSpring"                 = "Spring (v Winter)",
  "gph_tscore"                   = "Physical health",
  "gmh_tscore"                   = "Mental health",
  "cog_func_tscore"              = "Perceived cognition",
  "breq3_intrinsic_baseline_z"   = "Intrinsic motivation",
  "breq3_amotivation_baseline_z" = "Amotivation",
  "selfeff_total_score_baseli_z" = "Exercise self-efficacy"
)

or_forest <- bind_rows(
  or_tables$Opportunity %>% mutate(model = "Opportunity"),
  or_tables$Capability  %>% mutate(model = "Capability"),
  or_tables$Motivation  %>% mutate(model = "Motivation")
) %>%
  mutate(
    model        = factor(model, levels = c("Opportunity", "Capability", "Motivation")),
    term_display = dplyr::recode(term, !!!term_labels),
    term_display = reorder(term_display, estimate)
  )

# Stacked forest plot — one panel per domain model, all predictors shown.
# Points are coloured by direction (OR > 1 = positive, OR < 1 = negative).
ggplot(or_forest, aes(x = estimate, y = term_display, colour = estimate > 1)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, linewidth = 0.7) +
  geom_point(size = 2.8) +
  scale_x_log10() +
  scale_colour_manual(values = c("TRUE" = "#D55E00", "FALSE" = "#0072B2"),
                      guide  = "none") +
  facet_grid(model ~ ., scales = "free_y", space = "free_y") +
  labs(
    title    = "Odds Ratios",
    subtitle = "Outcome: Daily 30 minute Walk | error bars = 95% CI",
    x        = "Odds Ratio (log scale)",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
    strip.text    = element_text(face = "bold", size = 13),
    panel.spacing = unit(1, "lines")
  )


# ---- Predicted probability plots by focal predictor and group (type) ----
library(ggeffects)

# Helper: predicted probability ribbon plot from a glmer model.
# focal_term: ggpredict term string for the x-axis variable (e.g. "avg_temp_f [n=100]")
# x_label:    x-axis label for the plot
# title:      plot title
pred_prob_plot <- function(model, focal_term, x_label, title) {
  preds <- ggpredict(model,
                     terms = c(focal_term, "type"),
                     type  = "fixed")
  ggplot(preds, aes(x = x, y = predicted, color = group, fill = group)) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
    geom_line(linewidth = 1.2) +
    scale_color_manual(values = c("Control" = "#40B0A6", "Intervention" = "#D55E00")) +
    scale_fill_manual(values  = c("Control" = "#40B0A6", "Intervention" = "#D55E00")) +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    labs(
      title = title,
      x     = x_label,
      y     = "Predicted Probability of Walking ≥27 min",
      color = "Group", fill = "Group"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title      = element_text(hjust = 0.5, face = "bold"),
      legend.position = "bottom"
    )
}

# Intrinsic motivation (Motivation model)
pred_prob_plot(mod_motivation,
               "breq3_intrinsic_baseline_z [n=100]",
               "Intrinsic Motivation (z-scored)",
               "Predicted Probability of Walking by Intrinsic Motivation")

# Daily temperature (Opportunity model)
pred_prob_plot(mod_opportunity,
               "avg_temp_f [n=100]",
               "Daily Temperature (°F, within-person z-scored)",
               "Predicted Probability of Walking by Daily Temperature")

# Daily precipitation (Opportunity model)
pred_prob_plot(mod_opportunity,
               "precip_in [n=100]",
               "Daily Precipitation (mm, within-person z-scored)",
               "Predicted Probability of Walking by Daily Precipitation")

# Perceived cognition (Capability model)
pred_prob_plot(mod_capability,
               "cog_func_tscore [n=100]",
               "Perceived Cognition (T-score)",
               "Predicted Probability of Walking by Perceived Cognition")
```

### OR Tables

```{r}
library(kableExtra)

# Combine OR tables, apply publication labels, and add model column.
# tidy() returns log-scale estimates; re-extracting with exponentiate = FALSE
# gives the SE and z-stat on the log scale (correct), while ORs and CIs are
# exponentiated for display.
or_table_combined <- bind_rows(
  or_tables$Opportunity %>% mutate(model = "Opportunity"),
  or_tables$Capability  %>% mutate(model = "Capability"),
  or_tables$Motivation  %>% mutate(model = "Motivation")
) %>%
  mutate(
    term = dplyr::recode(term, !!!term_labels)
  ) %>%
  select(model, term, estimate, conf.low, conf.high, std.error, statistic) %>%
  mutate(
    across(c(estimate, conf.low, conf.high), ~ round(., 2)),
    std.error = round(std.error, 3),
    statistic = round(statistic, 2)
  ) %>%
  rename(
    Predictor  = term,
    OR         = estimate,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    SE         = std.error,
    `z`        = statistic
  )

# Row counts per model for pack_rows grouping
n_opp <- sum(or_table_combined$model == "Opportunity")
n_cap <- sum(or_table_combined$model == "Capability")
n_mot <- sum(or_table_combined$model == "Motivation")

or_table_combined %>%
  select(-model) %>%
  kable(
    caption = "Odds Ratios for Opportunity, Capability, and Motivation Models (Outcome: Daily 30-min Walk)",
    align   = c("l", "c", "c", "c", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 12
  ) %>%
  column_spec(1, bold = TRUE, width = "16em") %>%
  pack_rows("Opportunity", 1,          n_opp,                    bold = TRUE, italic = FALSE) %>%
  pack_rows("Capability",  n_opp + 1,  n_opp + n_cap,            bold = TRUE, italic = FALSE) %>%
  pack_rows("Motivation",  n_opp + n_cap + 1, n_opp + n_cap + n_mot,
            bold = TRUE, italic = FALSE)
```

### Models: has_completed_walk outcome

```{r}
fit_glmer_cw <- function(predictors, label) {
  f <- paste("has_completed_walk ~",
             paste(predictors, collapse = " + "),
             "+ (1 | participant_id)")
  cat(sprintf("\n=== %s model (has_completed_walk) ===\n", label))
  cat(sprintf("Formula: %s\n", f))
  tryCatch(
    glmer(as.formula(f), data = mods_final, family = binomial,
          control = glmerControl(optimizer = "bobyqa")),
    error = function(e) { cat("Model failed:", conditionMessage(e), "\n"); NULL }
  )
}

# --- Step 1: Demographics-only model ---
mod_demographics_cw <- fit_glmer_cw(demographics, "Demographics only")
cat("\n\n========== Demographics-only Model Summary (has_completed_walk) ==========\n")
print(summary(mod_demographics_cw))

# Extract significant demographic terms (p < .05), collapsing dummy suffixes
# back to the original variable name so whole factors are kept or dropped.
sig_demo_terms_cw <- tidy(mod_demographics_cw, effects = "fixed") %>%
  filter(term != "(Intercept)", p.value < 0.05) %>%
  pull(term)

sig_demographics_cw <- demographics[sapply(demographics,
                                           function(v) any(startsWith(sig_demo_terms_cw, v)))]

cat(sprintf("\nSignificant demographic(s) (p < .05): %s\n",
            paste(sig_demographics_cw, collapse = ", ")))
cat("These will be carried forward into the domain models.\n")

# --- Step 2: Domain models with only significant demographics ---
mod_opportunity_cw <- fit_glmer_cw(c(env_vars, sig_demographics_cw), "Opportunity")
mod_capability_cw  <- fit_glmer_cw(c(psy_vars, sig_demographics_cw),  "Capability")
mod_motivation_cw  <- fit_glmer_cw(c(mot_vars, sig_demographics_cw),  "Motivation")

# Summaries
for (nm in c("Opportunity", "Capability", "Motivation")) {
  m <- get(paste0("mod_", tolower(nm), "_cw"))
  cat(sprintf("\n\n========== %s Model Summary (has_completed_walk) ==========\n", nm))
  if (!is.null(m)) print(summary(m)) else cat("  Model unavailable.\n")
}

# Odds ratio tables
or_tables_cw <- lapply(
  list(Opportunity = mod_opportunity_cw,
       Capability  = mod_capability_cw,
       Motivation  = mod_motivation_cw),
  function(m) {
    if (is.null(m)) return(NULL)
    tidy(m, effects = "fixed", exponentiate = TRUE, conf.int = TRUE) %>%
      filter(term != "(Intercept)")
  }
)

cat("\n\n========== Opportunity OR Table (has_completed_walk) ==========\n"); print(or_tables_cw$Opportunity)
cat("\n========== Capability OR Table (has_completed_walk) ==========\n");  print(or_tables_cw$Capability)
cat("\n========== Motivation OR Table (has_completed_walk) ==========\n");  print(or_tables_cw$Motivation)
```

### Backward Elimination: has_completed_walk

```{r}
# Backward elimination for a single glmer model.
# At each iteration, the fixed effect with the largest p-value above `threshold`
# is dropped. Only one variable is removed per iteration. The loop stops when
# all remaining fixed effects have p <= threshold, or only one predictor remains.
#
# Note: p-values from glmer (Wald z) are approximate. They are used here purely
# as a guide for variable removal, not for formal inference on the final model.

backward_glmer <- function(predictors, outcome, label, threshold = 0.1) {

  current_preds <- predictors

  cat(sprintf("\n\n========== Backward Elimination: %s (%s) ==========\n",
              label, outcome))

  iteration <- 0

  repeat {
    iteration <- iteration + 1

    f <- paste(outcome, "~",
               paste(current_preds, collapse = " + "),
               "+ (1 | participant_id)")

    m <- tryCatch(
      glmer(as.formula(f), data = mods_final, family = binomial,
            control = glmerControl(optimizer = "bobyqa")),
      error = function(e) {
        cat("  Model failed:", conditionMessage(e), "\n")
        NULL
      }
    )

    if (is.null(m)) break

    # Extract fixed-effect p-values, excluding the intercept
    coef_tbl <- tidy(m, effects = "fixed") %>%
      filter(term != "(Intercept)") %>%
      arrange(desc(p.value))

    # Find the worst offender
    worst     <- coef_tbl$term[1]
    worst_p   <- coef_tbl$p.value[1]

    cat(sprintf("\n--- Iteration %d | %d predictor(s) ---\n", iteration, length(current_preds)))
    print(coef_tbl %>% select(term, estimate, std.error, statistic, p.value))

    if (worst_p <= threshold) {
      cat(sprintf("\n  All p-values <= %.2f. Stopping.\n", threshold))
      break
    }

    cat(sprintf("\n  Removing '%s' (p = %.4f)\n", worst, worst_p))

    # Remove the term from the predictor list.
    # strip dummy suffixes so we remove the whole factor at once
    # (e.g. "urban_typeSuburban" -> remove "urban_type")
    base_term <- current_preds[sapply(current_preds,
                                      function(v) startsWith(worst, v))]
    if (length(base_term) == 0) base_term <- worst
    current_preds <- setdiff(current_preds, base_term[1])

    if (length(current_preds) == 0) {
      cat("  No predictors remaining. Stopping.\n")
      break
    }
  }

  # Return the final model
  f_final <- paste(outcome, "~",
                   paste(current_preds, collapse = " + "),
                   "+ (1 | participant_id)")
  cat(sprintf("\n  Final formula: %s\n", f_final))

  final_m <- tryCatch(
    glmer(as.formula(f_final), data = mods_final, family = binomial,
          control = glmerControl(optimizer = "bobyqa")),
    error = function(e) { cat("Final model failed:", conditionMessage(e), "\n"); NULL }
  )

  final_m
}

# Run backward elimination for all three models
be_opportunity_cw <- backward_glmer(c(env_vars, demographics),
                                    "has_completed_walk", "Opportunity")

be_capability_cw  <- backward_glmer(c(psy_vars, demographics),
                                    "has_completed_walk", "Capability")

be_motivation_cw  <- backward_glmer(c(mot_vars, demographics),
                                    "has_completed_walk", "Motivation")

# OR tables for the final backward-eliminated models
or_tables_be_cw <- lapply(
  list(Opportunity = be_opportunity_cw,
       Capability  = be_capability_cw,
       Motivation  = be_motivation_cw),
  function(m) {
    if (is.null(m)) return(NULL)
    tidy(m, effects = "fixed", exponentiate = TRUE, conf.int = TRUE) %>%
      filter(term != "(Intercept)")
  }
)

cat("\n\n========== Opportunity OR Table — Backward Elimination (has_completed_walk) ==========\n")
print(or_tables_be_cw$Opportunity)
cat("\n========== Capability OR Table — Backward Elimination (has_completed_walk) ==========\n")
print(or_tables_be_cw$Capability)
cat("\n========== Motivation OR Table — Backward Elimination (has_completed_walk) ==========\n")
print(or_tables_be_cw$Motivation)
```

### Interaction: NatWalkInd_z × urban_type

Does walkability only predict daily walking in urban-dwelling participants?

```{r}
# Add NatWalkInd_z × urban_type interaction to the opportunity model
mod_opportunity_inter <- update(
  mod_opportunity_cw,
  . ~ . + NatWalkInd_z:precip_in
)

summary(mod_opportunity_inter)

# Likelihood ratio test: does the interaction improve model fit?
lrt_natwalk_urban <- anova(mod_opportunity_cw, mod_opportunity_inter, test = "Chisq")
print(lrt_natwalk_urban)

# OR table for the interaction model
tidy(mod_opportunity_inter, effects = "fixed", exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

# Plot predicted probabilities across the range of NatWalkInd_z, by urban_type
library(ggeffects)
ggpredict(mod_opportunity_inter, terms = c("NatWalkInd_z [all]", "urban_type")) |>
  plot() +
  labs(
    title    = "Predicted probability of walking ≥27 min",
    subtitle = "NatWalkInd_z × urban_type interaction",
    x        = "Neighborhood walkability (z)",
    y        = "P(walk ≥27 min)",
    colour   = "Urban type"
  )
```

does having more green space buffer the neg effect of precipitation

```{r}
# Add NatWalkInd_z × urban_type interaction to the opportunity model
mod_opportunity_inter <- update(
  mod_opportunity_cw,
  . ~ . + NatWalkInd_z:urban_type
)

summary(mod_opportunity_inter)

# Likelihood ratio test: does the interaction improve model fit?
lrt_natwalk_urban <- anova(mod_opportunity_cw, mod_opportunity_inter, test = "Chisq")
print(lrt_natwalk_urban)

# OR table for the interaction model
tidy(mod_opportunity_inter, effects = "fixed", exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

# Plot predicted probabilities across the range of NatWalkInd_z, by urban_type
library(ggeffects)
ggpredict(mod_opportunity_inter, terms = c("NatWalkInd_z [all]", "urban_type")) |>
  plot() +
  labs(
    title    = "Predicted probability of walking ≥27 min",
    subtitle = "NatWalkInd_z × urban_type interaction",
    x        = "Neighborhood walkability (z)",
    y        = "P(walk ≥27 min)",
    colour   = "Urban type"
  )
```

### Lasso: Variable Selection (has_completed_walk)

```{r}
library(tidymodels)

# All candidate predictors across the three theoretical domains
all_predictors <- c(env_vars, psy_vars, mot_vars, demographics)

# Prepare modelling data: keep only relevant columns, drop NAs
model_data <- mods_final %>%
  select(has_completed_walk, participant_id, all_of(all_predictors)) %>%
  drop_na()

# --- Recipe ---
# step_novel:    handles unseen factor levels
# step_dummy:    dummy-codes all nominal predictors (urban_type, season, sex_cat, etc.)
# step_zv:       removes zero-variance columns after dummy coding
# step_normalize: standardises all numeric predictors to mean=0, SD=1
#                 (required for lasso so penalty is applied on a common scale)
lasso_rec <- recipe(has_completed_walk ~ ., data = model_data) %>%
  update_role(participant_id, new_role = "ID") %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())
  #step_normalize(all_numeric_predictors())

# --- Leave-One-Participant-Out (LOPO) cross-validation ---
# v = number of unique participants → each fold holds out one participant entirely.
# Prevents same-participant observations appearing in both train and test sets,
# which would inflate performance estimates given the daily clustering structure.
lopo_folds <- group_vfold_cv(
  model_data,
  group = "participant_id",
  v     = n_distinct(model_data$participant_id)
)

# --- Model: logistic regression with lasso penalty ---
# mixture = 1 → pure lasso (variable selection)
# penalty is tuned via CV
lasso_spec <- logistic_reg(
  penalty = tune(),
  mixture = 1
) %>%
  set_engine("glmnet")

# --- Workflow ---
lasso_wf <- workflow() %>%
  add_recipe(lasso_rec) %>%
  add_model(lasso_spec)

# --- Penalty grid: 50 values log-spaced from 0.0001 to 1 ---
penalty_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  levels = 50
)

# --- Tune across LOPO folds ---
set.seed(42)
lasso_res <- tune_grid(
  lasso_wf,
  resamples = lopo_folds,
  grid      = penalty_grid,
  metrics   = metric_set(mn_log_loss),
  control   = control_grid(save_pred = TRUE)
)

# --- CV performance plot ---
autoplot(lasso_res)

# --- Select penalty using 1-SE rule ---
# Among all models within 1 SE of minimum log-loss, pick the one with the
# highest penalty (most regularised / fewest predictors).
# Appropriate for small N where CV estimates are noisy.
best_penalty <- select_by_one_std_err(lasso_res, metric = "mn_log_loss", desc(penalty))
cat(sprintf("Selected penalty (1-SE rule): %.5f
", best_penalty$penalty))

# Show CV metrics at selected penalty
lasso_res %>%
  collect_metrics() %>%
  filter(.metric == "mn_log_loss") %>%
  arrange(mean) %>%
  slice_head(n = 5) %>%
  print()

# --- Fit final model on full data at chosen penalty ---
final_wf  <- finalize_workflow(lasso_wf, best_penalty)
final_fit <- fit(final_wf, data = model_data)

# --- Extract selected predictors and odds ratios ---
# Only predictors with non-zero coefficients survived the lasso penalty.
# OR > 1: associated with higher daily probability of completing a walk.
# OR < 1: associated with lower probability.
# Absent predictors lacked sufficient signal given N=48.
coef_tbl <- tidy(final_fit) %>%
  filter(term != "(Intercept)", estimate != 0) %>%
  mutate(OR = exp(estimate)) %>%
  arrange(desc(abs(estimate)))

cat(sprintf("\n=== %d predictor(s) selected ===\n", nrow(coef_tbl)))
print(coef_tbl %>% select(term, estimate, OR))
```

### Lasso: Performance Metrics (has_completed_walk)

```{r}
# Held-out predictions at the selected penalty from LOPO-CV
# save_pred = TRUE means every fold's test predictions were stored
cv_preds <- collect_predictions(lasso_res, parameters = best_penalty)

# --- ROC-AUC ---
# Uses held-out probability estimates across all 48 LOPO folds.
# event_level = "second" because factor levels are c("False", "True")
roc_val <- cv_preds %>%
  roc_auc(truth = has_completed_walk, .pred_True, event_level = "second")

cat(sprintf("CV ROC-AUC: %.3f\n", roc_val$.estimate))

# --- CV log-loss at selected penalty ---
logloss_val <- lasso_res %>%
  collect_metrics() %>%
  filter(.metric == "mn_log_loss") %>%
  inner_join(best_penalty %>% select(penalty), by = "penalty")

cat(sprintf("CV log-loss: %.3f (SE: %.3f)\n",
            logloss_val$mean, logloss_val$std_err))

# --- Check predicted probability distribution ---
# If all probabilities are below 0.5, a fixed 0.5 cutoff will classify
# everything as "False" (sensitivity = 0, specificity = 1).
# This indicates class imbalance — use an outcome-informed threshold instead.
cat("\n--- Predicted probability summary (.pred_True) ---\n")
print(summary(cv_preds$.pred_True))

outcome_prev <- mean(model_data$has_completed_walk == "True")
cat(sprintf("Outcome prevalence (prop True): %.3f\n", outcome_prev))

# --- Optimal threshold via Youden's J (maximises sensitivity + specificity) ---
# More appropriate than 0.5 when classes are imbalanced.
roc_df <- cv_preds %>%
  roc_curve(truth = has_completed_walk, .pred_True, event_level = "second")

best_threshold <- roc_df %>%
  mutate(j_index = sensitivity + specificity - 1) %>%
  slice_max(j_index, n = 1) %>%
  pull(.threshold)

cat(sprintf("Optimal threshold (Youden's J): %.3f\n", best_threshold))

# --- Classification metrics at optimal threshold ---
cv_preds_class <- cv_preds %>%
  mutate(.pred_class = factor(
    ifelse(.pred_True >= best_threshold, "True", "False"),
    levels = c("False", "True")
  ))

class_metrics <- cv_preds_class %>%
  metric_set(accuracy, sens, spec)(
    truth       = has_completed_walk,
    estimate    = .pred_class,
    event_level = "second"
  )

cat("\n--- Classification metrics at Youden's J threshold ---\n")
print(class_metrics %>% select(.metric, .estimate))

# --- Confusion matrix ---
cat("\n--- Confusion matrix (LOPO-CV held-out predictions) ---\n")
cv_preds_class %>%
  conf_mat(truth = has_completed_walk, estimate = .pred_class) %>%
  print()

# --- ROC curve ---
cv_preds %>%
  roc_curve(truth = has_completed_walk, .pred_True, event_level = "second") %>%
  autoplot() +
  labs(
    title    = "ROC Curve — LOPO-CV held-out predictions",
    subtitle = "Lasso model | outcome: has_completed_walk",
    x        = "1 - Specificity",
    y        = "Sensitivity"
  )
```

### Elastic Net: Variable Selection (has_completed_walk)

```{r}
# Elastic net extends lasso by adding an L2 (ridge) penalty alongside the L1
# (lasso) penalty. The mixture parameter (alpha in glmnet) controls the blend:
#   mixture = 1 → pure lasso (sparse); mixture = 0 → pure ridge (all kept).
# Tuning mixture jointly with penalty allows the model to retain more predictors
# than lasso while still shrinking noisy coefficients toward zero.

# Reuse the same recipe and LOPO folds from the lasso section above.

# --- Model: logistic regression with elastic net penalty ---
# Both penalty (lambda) and mixture (alpha) are tuned.
enet_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet")

# --- Workflow ---
enet_wf <- workflow() %>%
  add_recipe(lasso_rec) %>%
  add_model(enet_spec)

# --- Tuning grid ---
# 10 penalty values (log-spaced 0.0001–1) × 5 mixture values (0.2–1.0)
# mixture values < 1 introduce the ridge component that keeps more predictors.
enet_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0.1, 1.0)),
  levels = c(penalty = 50, mixture = 5)
)

# --- Tune across LOPO folds ---
set.seed(42)
enet_res <- tune_grid(
  enet_wf,
  resamples = lopo_folds,
  grid      = enet_grid,
  metrics   = metric_set(mn_log_loss),
  control   = control_grid(save_pred = TRUE)
)

# --- CV performance plot ---
# Each line is one mixture value; x-axis is the penalty (log scale).
autoplot(enet_res)

# --- Select best hyperparameters using 1-SE rule ---
# Among all (penalty, mixture) combinations within 1 SE of minimum log-loss,
# choose the one with the highest penalty (most regularised / fewest predictors).
best_enet <- select_by_one_std_err(enet_res, metric = "mn_log_loss", desc(penalty))
cat(sprintf("Selected penalty: %.5f | Selected mixture: %.2f\n",
            best_enet$penalty, best_enet$mixture))

# Show top CV results
enet_res %>%
  collect_metrics() %>%
  filter(.metric == "mn_log_loss") %>%
  arrange(mean) %>%
  slice_head(n = 5) %>%
  print()

# --- Fit final model on full data at chosen hyperparameters ---
enet_final_wf  <- finalize_workflow(enet_wf, best_enet)
enet_final_fit <- fit(enet_final_wf, data = model_data)

# --- Extract selected predictors and odds ratios ---
# Predictors with non-zero coefficients survived the elastic net penalty.
enet_coef_tbl <- tidy(enet_final_fit) %>%
  filter(term != "(Intercept)", estimate != 0) %>%
  mutate(OR = exp(estimate)) %>%
  arrange(desc(abs(estimate)))

cat(sprintf("\n=== %d predictor(s) selected by elastic net ===\n", nrow(enet_coef_tbl)))
print(enet_coef_tbl %>% select(term, estimate, OR))
```

### Elastic Net: Performance Metrics (has_completed_walk)

```{r}
# Held-out predictions at the selected (penalty, mixture) from LOPO-CV
enet_cv_preds <- collect_predictions(enet_res, parameters = best_enet)

# --- ROC-AUC ---
enet_roc_val <- enet_cv_preds %>%
  roc_auc(truth = has_completed_walk, .pred_True, event_level = "second")

cat(sprintf("Elastic net CV ROC-AUC: %.3f\n", enet_roc_val$.estimate))

# --- CV log-loss at selected hyperparameters ---
enet_logloss_val <- enet_res %>%
  collect_metrics() %>%
  filter(.metric == "mn_log_loss") %>%
  inner_join(best_enet %>% select(penalty, mixture), by = c("penalty", "mixture"))

cat(sprintf("Elastic net CV log-loss: %.3f (SE: %.3f)\n",
            enet_logloss_val$mean, enet_logloss_val$std_err))

# --- Predicted probability summary ---
cat("\n--- Predicted probability summary (.pred_True) ---\n")
print(summary(enet_cv_preds$.pred_True))

cat(sprintf("Outcome prevalence (prop True): %.3f\n",
            mean(model_data$has_completed_walk == "True")))

# --- Optimal threshold via Youden's J ---
enet_roc_df <- enet_cv_preds %>%
  roc_curve(truth = has_completed_walk, .pred_True, event_level = "second")

enet_best_threshold <- enet_roc_df %>%
  mutate(j_index = sensitivity + specificity - 1) %>%
  slice_max(j_index, n = 1) %>%
  pull(.threshold)

cat(sprintf("Optimal threshold (Youden's J): %.3f\n", enet_best_threshold))

# --- Classification metrics at optimal threshold ---
enet_cv_preds_class <- enet_cv_preds %>%
  mutate(.pred_class = factor(
    ifelse(.pred_True >= enet_best_threshold, "True", "False"),
    levels = c("False", "True")
  ))

enet_class_metrics <- enet_cv_preds_class %>%
  metric_set(accuracy, sens, spec)(
    truth       = has_completed_walk,
    estimate    = .pred_class,
    event_level = "second"
  )

cat("\n--- Classification metrics at Youden's J threshold ---\n")
print(enet_class_metrics %>% select(.metric, .estimate))

# --- Confusion matrix ---
cat("\n--- Confusion matrix (LOPO-CV held-out predictions) ---\n")
enet_cv_preds_class %>%
  conf_mat(truth = has_completed_walk, estimate = .pred_class) %>%
  print()

# --- Lasso vs elastic net: side-by-side comparison ---
cat("\n--- Model comparison (LOPO-CV) ---\n")
comparison_tbl <- tibble(
  model        = c("Lasso", "Elastic Net"),
  penalty      = c(best_penalty$penalty, best_enet$penalty),
  mixture      = c(1.0,                  best_enet$mixture),
  cv_log_loss  = c(logloss_val$mean,     enet_logloss_val$mean),
  cv_roc_auc   = c(roc_val$.estimate,    enet_roc_val$.estimate),
  n_predictors = c(nrow(coef_tbl),       nrow(enet_coef_tbl))
)
print(comparison_tbl)

# --- ROC curve ---
enet_cv_preds %>%
  roc_curve(truth = has_completed_walk, .pred_True, event_level = "second") %>%
  autoplot() +
  labs(
    title    = "ROC Curve — LOPO-CV held-out predictions",
    subtitle = "Elastic net model | outcome: has_completed_walk",
    x        = "1 - Specificity",
    y        = "Sensitivity"
  )
```
------------------------------------------------------------------------

# Motivation & Self-Efficacy Outcomes: Intervention Effect and Mediation

**Purpose:** Two sets of models. First, the pre-post effect of the intervention versus control on all six BREQ-3 regulation subscales plus exercise self-efficacy (treating these as *outcomes*), reported with within-group simple slopes. Second, two simple mediation models asking whether pre-post change in amotivation or in intrinsic motivation mediates the intervention effect on total walks.

Note this is a different role for these variables than the `mod_motivation` model earlier in this document, which used *baseline* motivation as a predictor/modifier of daily walking.

```{r}
library(lme4)
library(lmerTest)
library(emmeans)
library(broom.mixed)

pal_grp <- c("Control" = "#0072B2", "Intervention" = "#D55E00")

# Seven outcomes: six BREQ-3 regulation subscales + exercise self-efficacy.
# Self-efficacy is on a different native scale from the BREQ-3 subscales, which is
# part of why all outcomes are standardised below.
outcome_specs <- tribble(
  ~measure,          ~v_baseline,                    ~v_endpoint,
  "Amotivation",     "breq3_amotivation_baseline",   "breq3_amotivation_endpoint",
  "External",        "breq3_external_baseline",      "breq3_external_endpoint",
  "Introjected",     "breq3_introjected_baseline",   "breq3_introjected_endpoint",
  "Identified",      "breq3_identified_baseline",    "breq3_identified_endpoint",
  "Integrated",      "breq3_integrated_baseline",    "breq3_integrated_endpoint",
  "Intrinsic",       "breq3_intrinsic_baseline",     "breq3_intrinsic_endpoint",
  "Self-efficacy",   "selfeff_total_score_baseli",   "selfeff_total_score_endpoi"
)
```

## Part 1 — Pre-post effect of intervention vs control (z-scored outcomes)

One linear mixed model per outcome: `score_z ~ group * time + (1 | id)`, where `time` is baseline vs endpoint. Two quantities are reported:

-   **Group × Time interaction** — the differential pre-post change in the intervention arm relative to control. This is the primary estimand.
-   **Simple slopes** — the within-group pre-post change (Endpoint − Baseline) *separately* for each arm, from the same fit. The interaction is exactly the difference between these two simple slopes.

Because seven outcomes are tested, p-values are additionally reported with Benjamini-Hochberg FDR correction.

### How the standardisation is done, and why it matters

Each outcome is z-scored **pooling both timepoints together** (one mean and one SD per measure, applied to baseline and endpoint alike). This is deliberate and is *not* the same as the `_z` columns built in the data-prep chunk at the top of this document, which scale each column separately.

Scaling baseline and endpoint separately would force each timepoint to have mean 0, which **removes the pre-post change entirely** — the very thing these models estimate. For example, intrinsic motivation rises 0.161 points raw, but after per-column scaling that shift becomes exactly 0. Pooled scaling preserves the change and simply re-expresses it in SD units of the measure. Because standardising an outcome is a linear rescaling, all test statistics and p-values are identical to the raw-score analysis; only the units of the estimates change (verified below).

```{r}
motiv_long <- outcome_specs %>%
  pmap_dfr(function(measure, v_baseline, v_endpoint) {
    completed %>%
      transmute(id, group,
                measure  = measure,
                Baseline = .data[[v_baseline]],
                Endpoint = .data[[v_endpoint]])
  }) %>%
  pivot_longer(c(Baseline, Endpoint), names_to = "time", values_to = "score_raw") %>%
  # Pooled z-scoring: one mean/SD per measure across BOTH timepoints, so the
  # pre-post shift is retained (see note above).
  group_by(measure) %>%
  mutate(score_z = as.numeric(scale(score_raw))) %>%
  ungroup() %>%
  mutate(time    = factor(time, levels = c("Baseline", "Endpoint")),
         group   = factor(group, levels = c("Control", "Intervention")),
         measure = factor(measure, levels = outcome_specs$measure))

cat("rows:", nrow(motiv_long), "| ids:", n_distinct(motiv_long$id),
    "| measures:", nlevels(motiv_long$measure), "\n")

# Observed cell means with 95% CIs (one-sample t intervals)
motiv_long %>%
  group_by(measure, time) %>%
  summarise(n = n(), mean_z = mean(score_z),
            se = sd(score_z) / sqrt(n()), .groups = "drop") %>%
  mutate(ci_low  = mean_z - qt(0.975, n - 1) * se,
         ci_high = mean_z + qt(0.975, n - 1) * se) %>%
  select(measure, time, n, mean_z, se, ci_low, ci_high) %>%
  kbl(digits = 3, caption = "Observed mean z-score by timepoint, with 95% CI") %>%
  kable_styling(full_width = FALSE)
```

```{r}
# Confirm the pooled scaling kept the pre-post shift that per-column scaling destroys.
# CI here is a paired interval on the within-person change, pooled across both arms.
motiv_long %>%
  select(id, measure, time, score_z) %>%
  pivot_wider(names_from = time, values_from = score_z) %>%
  mutate(change_z = Endpoint - Baseline) %>%
  group_by(measure) %>%
  summarise(n = n(), shift_z = mean(change_z),
            se = sd(change_z) / sqrt(n()), .groups = "drop") %>%
  mutate(ci_low  = shift_z - qt(0.975, n - 1) * se,
         ci_high = shift_z + qt(0.975, n - 1) * se,
         t_stat  = shift_z / se,
         p.value = 2 * pt(-abs(t_stat), df = n - 1)) %>%
  select(measure, n, shift_z, se, ci_low, ci_high, t_stat, p.value) %>%
  kbl(digits = 3,
      caption = "Overall pre-post shift in SD units with 95% CI (shift != 0 confirms change is retained)") %>%
  kable_styling(full_width = FALSE)
```

```{r}
fit_prepost <- function(msr) {
  d <- motiv_long %>% filter(measure == msr)
  m <- lmer(score_z ~ group * time + (1 | id), data = d)

  tid <- broom.mixed::tidy(m, effects = "fixed") %>%
    filter(term == "groupIntervention:timeEndpoint")

  # Simple slopes: within-group pre-post change, one per arm
  slopes <- contrast(emmeans(m, ~ time | group), "revpairwise", by = "group") %>%
    summary(infer = c(TRUE, TRUE)) %>%
    as.data.frame() %>%
    mutate(measure = msr)

  # Estimated marginal means at each group x time cell
  cells <- as.data.frame(emmeans(m, ~ time | group)) %>% mutate(measure = msr)

  list(model  = m,
       inter  = tibble(measure = msr, estimate = tid$estimate, std.error = tid$std.error,
                       df = tid$df, statistic = tid$statistic, p.value = tid$p.value),
       slopes = slopes,
       cells  = cells)
}

prepost <- map(levels(motiv_long$measure), fit_prepost) %>%
  set_names(levels(motiv_long$measure))

# CIs are t-based on the model's Satterthwaite df (not a normal +/- 1.96 approximation),
# so they are consistent with the emmeans-derived intervals in the tables below.
# NOTE: these intervals correspond to the UNADJUSTED p-values. They are not widened for
# the seven-outcome FDR correction, so a CI can exclude zero while p_fdr > .05.
inter_tbl <- map_dfr(prepost, "inter") %>%
  mutate(ci_low  = estimate - qt(0.975, df) * std.error,
         ci_high = estimate + qt(0.975, df) * std.error,
         p_fdr   = p.adjust(p.value, method = "BH"))

inter_tbl %>%
  select(measure, estimate, std.error, ci_low, ci_high, statistic, df, p.value, p_fdr) %>%
  kbl(digits = 3,
      caption = "Group x Time interaction (SD units) with 95% CI: differential pre-post change, Intervention vs Control") %>%
  kable_styling(full_width = FALSE)
```

### Simple slopes: within-group pre-post change

The within-group marginal effect of time, estimated separately in each arm. A slope whose CI excludes zero indicates that arm changed significantly from baseline to endpoint on its own.

```{r}
slopes_tbl <- map_dfr(prepost, "slopes") %>%
  select(measure, group, slope = estimate, SE, df,
         ci_low = lower.CL, ci_high = upper.CL, t.ratio, p.value) %>%
  mutate(measure = factor(measure, levels = outcome_specs$measure)) %>%
  arrange(measure, group)

slopes_tbl %>%
  kbl(digits = 3,
      caption = "Simple slopes (SD units) with 95% CI: Endpoint - Baseline within each arm") %>%
  kable_styling(full_width = FALSE) %>%
  pack_rows(index = table(slopes_tbl$measure))
```

```{r}
# Estimated marginal means at each group x time cell, with 95% CIs
map_dfr(prepost, "cells") %>%
  select(measure, group, time, emmean, SE, df,
         ci_low = lower.CL, ci_high = upper.CL) %>%
  mutate(measure = factor(measure, levels = outcome_specs$measure)) %>%
  arrange(measure, group, time) %>%
  kbl(digits = 3,
      caption = "Estimated marginal means (SD units) with 95% CI, by group and timepoint") %>%
  kable_styling(full_width = FALSE)
```

```{r}
# Verification 1: the interaction is exactly the difference in simple slopes
slope_diff_check <- slopes_tbl %>%
  select(measure, group, slope) %>%
  pivot_wider(names_from = group, values_from = slope) %>%
  mutate(diff_of_slopes = Intervention - Control) %>%
  left_join(inter_tbl %>% select(measure, interaction = estimate), by = "measure") %>%
  mutate(max_abs_discrepancy = abs(diff_of_slopes - interaction))
print(as.data.frame(slope_diff_check %>% mutate(across(where(is.numeric), ~round(., 8)))))

# Verification 2: z-scoring is a linear rescaling, so p-values match the raw-score models
raw_p <- map_dbl(levels(motiv_long$measure), function(msr) {
  d <- motiv_long %>% filter(measure == msr)
  broom.mixed::tidy(lmer(score_raw ~ group * time + (1 | id), data = d),
                    effects = "fixed") %>%
    filter(term == "groupIntervention:timeEndpoint") %>% pull(p.value)
})
cat("\nMax |p(z-scored) - p(raw)| across the seven outcomes:",
    max(abs(inter_tbl$p.value - raw_p)), "\n")
```

### Scatter plots: baseline vs endpoint by group

Each panel plots endpoint against baseline for one outcome, in SD units. The dashed 45-degree line is "no change": points above it increased, below it decreased. Group-specific regression lines sitting apart vertically indicate a differential pre-post effect (i.e. a non-zero interaction above).

```{r fig-breq-prepost, fig.width=10, fig.height=6}
motiv_wide <- motiv_long %>%
  select(id, group, measure, time, score_z) %>%
  pivot_wider(names_from = time, values_from = score_z)

ggplot(motiv_wide, aes(x = Baseline, y = Endpoint, color = group, fill = group)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.4) +
  geom_point(size = 1.8, alpha = 0.75) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              linewidth = 0.7, alpha = 0.15) +
  facet_wrap(~ measure, nrow = 2) +
  scale_color_manual(values = pal_grp) +
  scale_fill_manual(values = pal_grp) +
  labs(x = "Baseline (SD units)", y = "Endpoint (SD units)", color = NULL, fill = NULL,
       title = "Motivation and self-efficacy: baseline vs endpoint by group",
       subtitle = "Z-scored pooling both timepoints; dashed line = no change") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey94", color = NA),
        strip.text = element_text(face = "bold", size = 10))
```

### Simple slopes plot

The same within-group slopes as the table above, plotted so the two arms can be compared directly. Where the two intervals overlap heavily, the interaction is near zero.

```{r fig-simple-slopes, fig.width=8, fig.height=5}
ggplot(slopes_tbl, aes(x = slope, y = fct_rev(measure), color = group)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_pointrange(aes(xmin = ci_low, xmax = ci_high),
                  position = position_dodge(width = 0.55),
                  size = 0.45, linewidth = 0.6) +
  scale_color_manual(values = pal_grp) +
  labs(x = "Within-group pre-post change (SD units, Endpoint - Baseline)",
       y = NULL, color = NULL,
       title = "Simple slopes by arm",
       subtitle = "Intervals excluding zero indicate significant within-arm change") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Part 2 — Mediation: does change in amotivation or intrinsic motivation mediate the effect on total walks?

Two simple mediation models, `X -> M -> Y`:

-   **X** = group (Control = 0, Intervention = 1)
-   **M** = pre-post change in the mediator (endpoint − baseline), one model for amotivation and one for intrinsic motivation
-   **Y** = total walks, weeks 1-12 (person-level count)

Each model adjusts for the **baseline value of the mediator** in both the a-path and the b-path. Change scores are correlated with their own baseline (strongly here, given floor/ceiling effects), so leaving baseline out would confound both paths.

Inference on the indirect effect uses a **nonparametric bootstrap percentile CI** (5,000 resamples) rather than a normal-theory test, because the product `a*b` is not normally distributed. The bootstrap also protects the inference against the heteroscedasticity of the raw count outcome.

```{r}
walks_person <- weekly_walks %>%
  filter(week <= 12) %>%
  group_by(id) %>%
  summarise(total_walks = sum(weekly_walks, na.rm = TRUE), .groups = "drop")

med_dat <- completed %>%
  select(id, group,
         amot_pre = breq3_amotivation_baseline, amot_post = breq3_amotivation_endpoint,
         intr_pre = breq3_intrinsic_baseline,   intr_post = breq3_intrinsic_endpoint) %>%
  left_join(walks_person, by = "id") %>%
  mutate(X      = as.integer(factor(group, levels = c("Control", "Intervention"))) - 1L,
         d_amot = amot_post - amot_pre,
         d_intr = intr_post - intr_pre) %>%
  filter(!is.na(total_walks), !is.na(d_amot), !is.na(d_intr))

cat("Mediation analytic n:", nrow(med_dat), "\n")

med_dat %>%
  group_by(group) %>%
  summarise(n = n(),
            mean_change_amot = mean(d_amot), mean_change_intr = mean(d_intr),
            mean_total_walks = mean(total_walks), .groups = "drop") %>%
  kbl(digits = 3, caption = "Mediator change and outcome by group") %>%
  kable_styling(full_width = FALSE)
```

```{r}
# a-path: M ~ X + baseline(M)
# b-path: Y ~ M + X + baseline(M)   (b = mediator effect; X coefficient = direct effect)
# total:  Y ~ X + baseline(M)
run_mediation <- function(dat, m_var, m_pre, y_var = "total_walks",
                          nboot = 5000, seed = 42) {
  f_a <- reformulate(c("X", m_pre),        response = m_var)
  f_b <- reformulate(c(m_var, "X", m_pre), response = y_var)
  f_c <- reformulate(c("X", m_pre),        response = y_var)

  est <- function(d) {
    a  <- coef(lm(f_a, data = d))[["X"]]
    mb <- lm(f_b, data = d)
    b  <- coef(mb)[[m_var]]
    cp <- coef(mb)[["X"]]
    c(a = a, b = b, indirect = a * b, direct = cp,
      total = coef(lm(f_c, data = d))[["X"]])
  }

  point <- est(dat)
  set.seed(seed)
  boot <- replicate(nboot, {
    d <- dat[sample(nrow(dat), replace = TRUE), ]
    tryCatch(est(d), error = function(e) rep(NA_real_, 5))
  })
  ci <- apply(boot, 1, quantile, probs = c(0.025, 0.975), na.rm = TRUE)

  tibble(term = names(point), estimate = point,
         boot_se = apply(boot, 1, sd, na.rm = TRUE),
         ci_low = ci[1, ], ci_high = ci[2, ]) %>%
    mutate(crosses_zero = ci_low <= 0 & ci_high >= 0)
}

med_amot <- run_mediation(med_dat, "d_amot", "amot_pre")
med_intr <- run_mediation(med_dat, "d_intr", "intr_pre")

bind_rows(
  med_amot %>% mutate(mediator = "Change in amotivation"),
  med_intr %>% mutate(mediator = "Change in intrinsic motivation")
) %>%
  select(mediator, term, estimate, boot_se, ci_low, ci_high, crosses_zero) %>%
  kbl(digits = 3,
      caption = "Mediation results (bootstrap percentile CIs, 5000 resamples). Outcome = total walks, weeks 1-12") %>%
  kable_styling(full_width = FALSE) %>%
  pack_rows(index = c("Change in amotivation" = 5, "Change in intrinsic motivation" = 5))
```

```{r}
# Component models behind the two mediations
for (spec in list(c("d_amot", "amot_pre", "Amotivation"),
                  c("d_intr", "intr_pre", "Intrinsic motivation"))) {
  cat("\n========== ", spec[3], " ==========\n", sep = "")
  cat("--- a-path ---\n")
  print(round(summary(lm(reformulate(c("X", spec[2]), response = spec[1]),
                         data = med_dat))$coefficients, 4))
  cat("--- b-path (X coefficient here is the direct effect) ---\n")
  print(round(summary(lm(reformulate(c(spec[1], "X", spec[2]), response = "total_walks"),
                         data = med_dat))$coefficients, 4))
}
```

### b-path added-variable plots

These are **added-variable (partial residual) plots**, not raw scatters: both axes are residualised on group and baseline mediator, so the slope of each fitted line is exactly the `b` coefficient from the mediation model (verified in the output below). A raw scatter of mediator change against total walks is misleading here, because group is by far the dominant driver of walks and the reported `b` is a *partial* coefficient.

```{r fig-breq-bpath, fig.width=8, fig.height=4}
av_data <- function(dat, m_var, m_pre, label) {
  ry <- resid(lm(reformulate(c("X", m_pre), response = "total_walks"), data = dat))
  rm <- resid(lm(reformulate(c("X", m_pre), response = m_var),         data = dat))
  b  <- coef(lm(reformulate(c(m_var, "X", m_pre), response = "total_walks"),
                data = dat))[[m_var]]
  cat(label, "— added-variable slope:", round(coef(lm(ry ~ rm))[2], 4),
      "| model b:", round(b, 4), "\n")
  tibble(mediator = label, m_resid = rm, y_resid = ry, group = dat$group)
}

av <- bind_rows(
  av_data(med_dat, "d_amot", "amot_pre", "Amotivation"),
  av_data(med_dat, "d_intr", "intr_pre", "Intrinsic")
)

ggplot(av, aes(x = m_resid, y = y_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70", linewidth = 0.4) +
  geom_point(aes(color = group), size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "black", linewidth = 0.7, alpha = 0.15) +
  facet_wrap(~ mediator, nrow = 1, scales = "free_x") +
  scale_color_manual(values = pal_grp) +
  labs(x = "Mediator change | group, baseline (residual)",
       y = "Total walks | group, baseline (residual)", color = NULL,
       title = "b-path added-variable plots",
       subtitle = "Slope of each fitted line equals the b coefficient in the mediation model") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey94", color = NA),
        strip.text = element_text(face = "bold", size = 10))
```
