---
title: "Roybal - Outcomes"
format: html
---

# Roybal Final Data Analysis — Outcomes

**Purpose:** This document covers exploratory outcome analyses of cogntive outcomes from the Roybal pilot RCT.

### Load data

```{r}
library(emmeans)
library(dplyr)

# Loads: completed, dropped, all, weekly_walks, all_walks, weather_raw
load("roybal_base.RData")

# Composite variables built from columns already in completed (masterdoc sheet 1)
# Verbal Fluency
completed$baseline_vf_total <- completed$verbal_fluency_f + completed$verbal_fluency_a +
                                completed$verbal_fluency_s + completed$verbal_fluency_animals
completed$endpoint_vf_total <- completed$verbal_fluency_c + completed$verbal_fluency_f2 +
                                completed$verbal_fluency_l + completed$verbal_fluency_fruits

# HVLT
completed$hvlt_baseline_total_recall   <- completed$hvlt_baseline_trial1 + completed$hvlt_baseline_trial2 + completed$hvlt_baseline_trial3
completed$hvlt_baseline_delayed_recall <- completed$hvlt_baseline_trial4
completed$hvlt_baseline_retention      <- (completed$hvlt_baseline_trial4 / completed$hvlt_baseline_trial3) * 100
completed$hvlt_baseline_rdi            <- completed$discriminationindexdelay_form1

completed$hvlt_endpoint_total_recall   <- completed$hvlt_endpoint_trial1 + completed$hvlt_endpoint_trial2 + completed$hvlt_endpoint_trial3
completed$hvlt_endpoint_delayed_recall <- completed$hvlt_endpoint_trial4
completed$hvlt_endpoint_retention      <- (completed$hvlt_endpoint_trial4 / completed$hvlt_endpoint_trial3) * 100
completed$hvlt_endpoint_rdi            <- completed$discrimination_indexdelay_form2
```

### Group generation

- This creates a variable called group_who that dichotomies participants into whether on average (for at least 6 of 12 weeks) walked for 150 minutes or more

```{r}
# create variable that groups participants by whether ot not they walked for 150s on average for at least 6 of the 12 weeks. 
completed <- weekly_walks %>%
  group_by(id) %>%
  summarise(group_who = as.integer(sum(total_duration_walks > 150) >= 6)) %>%
  left_join(completed, ., by = "id")

# ---- Stacked proportion plot: group_who by study condition ----
completed %>%
  filter(!is.na(group_who), !is.na(group)) %>%
  mutate(
    group_who_label = factor(group_who, levels = c(0, 1),
                             labels = c("Did Not Meet", "Met Threshold"))
  ) %>%
  count(group_who_label, group) %>%
  group_by(group_who_label) %>%
  mutate(prop = n / sum(n),
         label = paste0(round(prop * 100), "%")) %>%
  ungroup() %>%
  ggplot(aes(x = group_who_label, y = n, fill = group)) +
  geom_col(position = "stack") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 4, color = "black") +
  scale_fill_manual(values = c("Control" = "#40B0A6", "Intervention" = "#FFBE6A")) +
  scale_y_continuous(breaks = scales::pretty_breaks()) +
  labs(
    title = "150 mins of walking per week",
    x     = "Walking Threshold Group",
    y     = "Number of Participants",
    fill  = "Condition"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )
```

### Adherence depth: how many weeks above 150 min?

```{r}
library(forcats)
library(kableExtra)

# per-participant: total weeks > 150, derived group_who, additional weeks beyond threshold
adherence_summary <- weekly_walks %>%
  filter(week <= 12) %>%
  group_by(id) %>%
  summarise(
    n_weeks_above = sum(total_duration_walks > 150, na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  mutate(
    group_who        = as.integer(n_weeks_above >= 6),
    additional_weeks = if_else(group_who == 1L, n_weeks_above - 6L, NA_integer_)
  )

# join back to week-level data
weekly_above <- weekly_walks %>%
  filter(week <= 12) %>%
  mutate(above_150 = total_duration_walks > 150) %>%
  left_join(adherence_summary, by = "id")

# ── Plot 1: tile heatmap — participant × week ─────────────────────────────────
weekly_above %>%
  mutate(
    id_sorted   = fct_reorder(id, n_weeks_above, .desc = TRUE),
    group_label = factor(group_who, levels = c(1, 0),
                         labels = c("Met threshold (≥6 wks)", "Did not meet (<6 wks)"))
  ) %>%
  ggplot(aes(x = week, y = id_sorted, fill = above_150)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_vline(xintercept = 6.5, linetype = "dashed", color = "black", linewidth = 0.7) +
  facet_grid(group_label ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values   = c("TRUE" = "#2196F3", "FALSE" = "#E0E0E0"),
    labels   = c("TRUE" = "> 150 min", "FALSE" = "≤ 150 min"),
    na.value = "grey90"
  ) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title    = "Weekly walking duration > 150 min — by participant and week",
    subtitle = "Dashed line at week 6 (minimum threshold); sorted by total weeks above threshold",
    x        = "Week",
    y        = "Participant",
    fill     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y     = element_text(size = 7),
    strip.text.y    = element_text(angle = 0, face = "bold"),
    legend.position = "bottom"
  )

# ── Plot 2: distribution of total weeks > 150 among group 1 ──────────────────
adherence_summary %>%
  filter(group_who == 1) %>%
  count(n_weeks_above) %>%
  ggplot(aes(x = factor(n_weeks_above), y = n)) +
  geom_col(fill = "#2196F3", width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4, size = 4.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Group 1: distribution of total weeks with > 150 min walking",
    subtitle = "Threshold = 6 weeks; x-axis shows total weeks achieved",
    x        = "Total weeks with > 150 min",
    y        = "Number of participants"
  ) +
  theme_minimal(base_size = 13)

# ── Table: group 1 — total and additional weeks beyond threshold ──────────────
adherence_summary %>%
  filter(group_who == 1) %>%
  arrange(desc(n_weeks_above)) %>%
  select(id, n_weeks_above, additional_weeks) %>%
  rename(
    Participant        = id,
    `Total weeks >150` = n_weeks_above,
    `Additional weeks` = additional_weeks
  ) %>%
  kbl(
    caption  = "Group 1 participants: total and additional weeks with > 150 min walking",
    booktabs = TRUE,
    align    = "lcc"
  ) %>%
  kable_styling(full_width = FALSE)
```

### PROMIS

##### Cognitive health

```{r}
# PROMIS Cognitive
promis_cog_long <- completed %>%
  select(id, group, group_who, age2, sex_cat, years_of_ed, days_since_injury, cog_func_tscore, cog_func_tscore_v2) %>%
  pivot_longer(cols = c(cog_func_tscore, cog_func_tscore_v2), names_to = "timepoint", values_to = "score") %>%
  mutate(timepoint_numeric = ifelse(timepoint == "cog_func_tscore", 0, 1))

# Model
m1_promis_cog <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = promis_cog_long)
summary(m1_promis_cog)

# Omnibus visual check (panel of all assumption plots) ---
result <- check_model(m1_promis_cog)
plot(result, panel = FALSE)
check_outliers(m1_promis_cog)

# Remove the single influential observation (row 71)
model_data_clean <- promis_cog_long[-71, ]

# Refit the model on the cleaned data
model_clean <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = model_data_clean)
summary(model_clean)

# Compare influential-case-removed model to original
compare_performance(m1_promis_cog, model_clean)

# Recheck outliers to confirm resolved
check_outliers(model_clean)

# Remove the single influential observation (row 71)
model_data_clean_2 <- promis_cog_long[-70, ]

# Refit the model on the cleaned data
model_clean_2 <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = model_data_clean_2)
summary(model_clean_2)

# Compare influential-case-removed model to original
compare_performance(m1_promis_cog, model_clean, model_clean_2)

  # asusmptions check for final model
result <- check_model(model_clean_2)
plot(result, panel = FALSE)

# standardized ES
standardize_parameters(model_clean_2, method = "refit") %>%
  filter(Parameter == "group_who:timepoint_numeric")

means_df <- model_data_clean_2 %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  ) %>%
  group_by(Group, TimePoint) %>%
  summarise(
    n    = n(),
    Mean = mean(score, na.rm = TRUE),
    SE   = sd(score, na.rm = TRUE) / sqrt(n),
    .groups = "drop"
  )

individual_df <- model_data_clean_2 %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  )

# Marginal means at pre/post for each group_who level
emm <- emmeans(model_clean_2, ~ group_who * timepoint_numeric,
               at = list(timepoint_numeric = c(0, 1)))
emm

# 1. group_who=1 vs group_who=0 AT POST-TEST (timepoint=1)
contrast(emm, method = "revpairwise", by = "timepoint_numeric")

# 2. Pre-to-post change WITHIN each group_who level (marginal effects)
contrast(emm, method = "revpairwise", by = "group_who")

# PLOT

ggplot() +
  # Individual trajectories
  geom_line(data = individual_df,
            aes(x = TimePoint, y = score, group = id, color = Group),
            alpha = 0.15, position = position_dodge(width = 0.2)) +
  geom_point(data = individual_df,
             aes(x = TimePoint, y = score, group = id, color = Group),
             alpha = 0.3, size = 2, position = position_dodge(width = 0.2)) +
  # Group mean lines
  geom_line(data = means_df,
            aes(x = TimePoint, y = Mean, group = Group, color = Group),
            linewidth = 1.5) +
  geom_point(data = means_df,
             aes(x = TimePoint, y = Mean, group = Group, color = Group),
             size = 4) +
  geom_errorbar(data = means_df,
                aes(x = TimePoint, ymin = Mean - SE, ymax = Mean + SE, group = Group, color = Group),
                width = 0.1, linewidth = 1) +
  scale_color_manual(values = c("Did Not Meet" = "#40B0A6", "Met Threshold" = "#FFBE6A")) +
  labs(
    title    = "Cognitive Function T-Score (Higher scores are better)",
    subtitle = "Pre-Post Changes by Walking Threshold Group",
    x        = "Assessment Time",
    y        = "Score",
    color    = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title      = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle   = element_text(hjust = 0.5)
  )
```

##### Physical health

```{r}
# PROMIS Physical Health
promis_phys_long <- completed %>%
  select(id, group, group_who, age2, sex_cat, years_of_ed, days_since_injury, gph_tscore, gph_tscore_v2) %>%
  pivot_longer(cols = c(gph_tscore, gph_tscore_v2), names_to = "timepoint", values_to = "score") %>%
  mutate(timepoint_numeric = ifelse(timepoint == "gph_tscore", 0, 1))

# model 
m1_promis_phys <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = promis_phys_long)
summary(m1_promis_phys)

# Recheck outliers to confirm resolved
check_outliers(m1_promis_phys)

# standardized ES
standardize_parameters(m1_promis_phys, method = "refit") %>%
  filter(Parameter == "group_who:timepoint_numeric")
```

##### Mental Health

```{r}
# PROMIS Mental Health
promis_mental_long <- completed %>%
  select(id, group, age2, group_who, sex_cat, years_of_ed, days_since_injury, gmh_tscore, gmh_tscore_v2) %>%
  pivot_longer(cols = c(gmh_tscore, gmh_tscore_v2), names_to = "timepoint", values_to = "score") %>%
  mutate(timepoint_numeric = ifelse(timepoint == "gmh_tscore", 0, 1))

# model 
m1_promis_men <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = promis_phys_long)
summary(m1_promis_men)

# Recheck outliers to confirm resolved
check_outliers(m1_promis_men)

# standardized ES
standardize_parameters(m1_promis_men, method = "refit") %>%
  filter(Parameter == "group_who:timepoint_numeric")
```

##### FDR corrections

```{r}
  # FDR corrections 
   FDR correction on interaction term p-values
interaction_pvals <- c(
  m1_promis_men   = coef(summary(m1_promis_men))["group_who.x:timepoint_numeric", "Pr(>|t|)"],
  m1_promis_phys  = coef(summary(m1_promis_phys))["group_who.x:timepoint_numeric", "Pr(>|t|)"],
  model_clean_2   = coef(summary(model_clean_2))["group_who.x:timepoint_numeric", "Pr(>|t|)"]
)

fdr_adjusted <- p.adjust(interaction_pvals, method = "fdr")

data.frame(
  model       = names(interaction_pvals),
  p_raw       = interaction_pvals,
  p_fdr       = fdr_adjusted,
  sig_fdr     = fdr_adjusted < 0.05
)
```

### Cognition

##### prepare data

```{r}
# Prepare long format data including group_who.x
vf_long_who <- completed %>%
  select(id, group, group_who, age2, sex_cat, years_of_ed, days_since_injury, baseline_vf_total, endpoint_vf_total) %>%
  pivot_longer(cols = c(baseline_vf_total, endpoint_vf_total), names_to = "timepoint", values_to = "score") %>%
  mutate(timepoint_numeric = ifelse(timepoint == "baseline_vf_total", 0, 1))

hvlt_total_long_who <- completed %>%
  select(id, group, group_who, age2, sex_cat, years_of_ed, days_since_injury, hvlt_baseline_total_recall, hvlt_endpoint_total_recall) %>%
  pivot_longer(cols = c(hvlt_baseline_total_recall, hvlt_endpoint_total_recall), names_to = "timepoint", values_to = "score") %>%
  mutate(timepoint_numeric = ifelse(timepoint == "hvlt_baseline_total_recall", 0, 1))

hvlt_delayed_long_who <- completed %>%
  select(id, group, group_who, age2, sex_cat, years_of_ed, days_since_injury, hvlt_baseline_delayed_recall, hvlt_endpoint_delayed_recall) %>%
  pivot_longer(cols = c(hvlt_baseline_delayed_recall, hvlt_endpoint_delayed_recall), names_to = "timepoint", values_to = "score") %>%
  mutate(timepoint_numeric = ifelse(timepoint == "hvlt_baseline_delayed_recall", 0, 1))

hvlt_retention_long_who <- completed %>%
  select(id, group, group_who, age2, sex_cat, years_of_ed, days_since_injury, hvlt_baseline_retention, hvlt_endpoint_retention) %>%
  pivot_longer(cols = c(hvlt_baseline_retention, hvlt_endpoint_retention), names_to = "timepoint", values_to = "score") %>%
  mutate(timepoint_numeric = ifelse(timepoint == "hvlt_baseline_retention", 0, 1))

```

##### Verbal fluency

```{r}

# Verbal Fluency
m1_vf_who <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = vf_long_who)
summary(m1_vf_who)
check_outliers(m1_vf_who)

# standardized ES
standardize_parameters(m1_vf_who, method = "refit") %>%
  filter(Parameter == "group_who:timepoint_numeric")

means_df_vf <- vf_long_who %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  ) %>%
  group_by(Group, TimePoint) %>%
  summarise(n = n(), Mean = mean(score, na.rm = TRUE), SE = sd(score, na.rm = TRUE) / sqrt(n), .groups = "drop")

individual_df_vf <- vf_long_who %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  )

ggplot() +
  geom_line(data = individual_df_vf,
            aes(x = TimePoint, y = score, group = id, color = Group),
            alpha = 0.15, position = position_dodge(width = 0.2)) +
  geom_point(data = individual_df_vf,
             aes(x = TimePoint, y = score, group = id, color = Group),
             alpha = 0.3, size = 2, position = position_dodge(width = 0.2)) +
  geom_line(data = means_df_vf,
            aes(x = TimePoint, y = Mean, group = Group, color = Group),
            linewidth = 1.5) +
  geom_point(data = means_df_vf,
             aes(x = TimePoint, y = Mean, group = Group, color = Group),
             size = 4) +
  geom_errorbar(data = means_df_vf,
                aes(x = TimePoint, ymin = Mean - SE, ymax = Mean + SE, group = Group, color = Group),
                width = 0.1, linewidth = 1) +
  scale_color_manual(values = c("Did Not Meet" = "#40B0A6", "Met Threshold" = "#FFBE6A")) +
  labs(
    title    = "Verbal Fluency (Higher scores are better)",
    subtitle = "Pre-Post Changes by Walking Threshold Group",
    x        = "Assessment Time",
    y        = "Score",
    color    = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

##### HVLT total recall

```{r}
# HVLT Total Recall
m1_hvlt_total_who <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = hvlt_total_long_who)
summary(m1_hvlt_total_who)
check_outliers(m1_hvlt_total_who)

# standardized ES
standardize_parameters(m1_hvlt_total_who, method = "refit") %>%
  filter(Parameter == "group_who:timepoint_numeric")

means_df_hvlt_total <- hvlt_total_long_who %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  ) %>%
  group_by(Group, TimePoint) %>%
  summarise(n = n(), Mean = mean(score, na.rm = TRUE), SE = sd(score, na.rm = TRUE) / sqrt(n), .groups = "drop")

individual_df_hvlt_total <- hvlt_total_long_who %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  )

ggplot() +
  geom_line(data = individual_df_hvlt_total,
            aes(x = TimePoint, y = score, group = id, color = Group),
            alpha = 0.15, position = position_dodge(width = 0.2)) +
  geom_point(data = individual_df_hvlt_total,
             aes(x = TimePoint, y = score, group = id, color = Group),
             alpha = 0.3, size = 2, position = position_dodge(width = 0.2)) +
  geom_line(data = means_df_hvlt_total,
            aes(x = TimePoint, y = Mean, group = Group, color = Group),
            linewidth = 1.5) +
  geom_point(data = means_df_hvlt_total,
             aes(x = TimePoint, y = Mean, group = Group, color = Group),
             size = 4) +
  geom_errorbar(data = means_df_hvlt_total,
                aes(x = TimePoint, ymin = Mean - SE, ymax = Mean + SE, group = Group, color = Group),
                width = 0.1, linewidth = 1) +
  scale_color_manual(values = c("Did Not Meet" = "#40B0A6", "Met Threshold" = "#FFBE6A")) +
  labs(
    title    = "HVLT Total Recall (Higher scores are better)",
    subtitle = "Pre-Post Changes by Walking Threshold Group",
    x        = "Assessment Time",
    y        = "Score",
    color    = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

##### HVLT delayed recall

```{r}

# HVLT Delayed Recall - primary model
m1_hvlt_delayed_who <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = hvlt_delayed_long_who)
summary(m1_hvlt_delayed_who)

# Detect outliers
outliers_hvlt_delayed <- check_outliers(m1_hvlt_delayed_who)
outliers_hvlt_delayed
outlier_rows_hvlt_delayed <- which(outliers_hvlt_delayed)
cat("Outlier rows:", outlier_rows_hvlt_delayed, "\n")

# Remove outliers and refit
hvlt_delayed_clean <- hvlt_delayed_long_who[-outlier_rows_hvlt_delayed, ]
m1_hvlt_delayed_clean <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = hvlt_delayed_clean)
summary(m1_hvlt_delayed_clean)

# Compare original vs cleaned model
compare_performance(m1_hvlt_delayed_who, m1_hvlt_delayed_clean)

# Recheck outliers on cleaned model - second round
outliers_hvlt_delayed_2 <- check_outliers(m1_hvlt_delayed_clean)
outliers_hvlt_delayed_2
outlier_rows_hvlt_delayed_2 <- which(outliers_hvlt_delayed_2)
cat("Second-round outlier rows:", outlier_rows_hvlt_delayed_2, "\n")

# Remove second outlier and refit
hvlt_delayed_clean_2 <- hvlt_delayed_clean[-outlier_rows_hvlt_delayed_2, ]
m1_hvlt_delayed_clean_2 <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = hvlt_delayed_clean_2)
summary(m1_hvlt_delayed_clean_2)

# Compare all three models
compare_performance(m1_hvlt_delayed_who, m1_hvlt_delayed_clean, m1_hvlt_delayed_clean_2)

# Recheck outliers - third round
outliers_hvlt_delayed_3 <- check_outliers(m1_hvlt_delayed_clean_2)
outliers_hvlt_delayed_3
outlier_rows_hvlt_delayed_3 <- which(outliers_hvlt_delayed_3)
cat("Third-round outlier rows:", outlier_rows_hvlt_delayed_3, "\n")

# Remove third outlier and refit
hvlt_delayed_clean_3 <- hvlt_delayed_clean_2[-outlier_rows_hvlt_delayed_3, ]
m1_hvlt_delayed_clean_3 <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = hvlt_delayed_clean_3)
summary(m1_hvlt_delayed_clean_3)

# Compare all four models
compare_performance(m1_hvlt_delayed_who, m1_hvlt_delayed_clean, m1_hvlt_delayed_clean_2, m1_hvlt_delayed_clean_3)

# Recheck outliers to confirm resolved
check_outliers(m1_hvlt_delayed_clean_3)

# Assumptions check on final model
result_delayed <- check_model(m1_hvlt_delayed_clean_3)
plot(result_delayed, panel = FALSE)

# standardized ES
standardize_parameters(m1_hvlt_delayed_clean_3, method = "refit") %>%
  filter(Parameter == "group_who:timepoint_numeric")

means_df_hvlt_delayed <- hvlt_delayed_clean_3 %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  ) %>%
  group_by(Group, TimePoint) %>%
  summarise(n = n(), Mean = mean(score, na.rm = TRUE), SE = sd(score, na.rm = TRUE) / sqrt(n), .groups = "drop")

individual_df_hvlt_delayed <- hvlt_delayed_clean_3 %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  )

ggplot() +
  geom_line(data = individual_df_hvlt_delayed,
            aes(x = TimePoint, y = score, group = id, color = Group),
            alpha = 0.15, position = position_dodge(width = 0.2)) +
  geom_point(data = individual_df_hvlt_delayed,
             aes(x = TimePoint, y = score, group = id, color = Group),
             alpha = 0.3, size = 2, position = position_dodge(width = 0.2)) +
  geom_line(data = means_df_hvlt_delayed,
            aes(x = TimePoint, y = Mean, group = Group, color = Group),
            linewidth = 1.5) +
  geom_point(data = means_df_hvlt_delayed,
             aes(x = TimePoint, y = Mean, group = Group, color = Group),
             size = 4) +
  geom_errorbar(data = means_df_hvlt_delayed,
                aes(x = TimePoint, ymin = Mean - SE, ymax = Mean + SE, group = Group, color = Group),
                width = 0.1, linewidth = 1) +
  scale_color_manual(values = c("Did Not Meet" = "#40B0A6", "Met Threshold" = "#FFBE6A")) +
  labs(
    title    = "HVLT Delayed Recall (Higher scores are better)",
    subtitle = "Pre-Post Changes by Walking Threshold Group",
    x        = "Assessment Time",
    y        = "Score",
    color    = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

##### HVLT retention

```{r}

# HVLT Retention - primary model
m1_hvlt_retention_who <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = hvlt_retention_long_who)
summary(m1_hvlt_retention_who)

# Detect outliers
outliers_hvlt_retention <- check_outliers(m1_hvlt_retention_who)
outliers_hvlt_retention
outlier_rows_hvlt_retention <- which(outliers_hvlt_retention)
cat("Outlier rows:", outlier_rows_hvlt_retention, "\n")

# Remove outliers and refit
hvlt_retention_clean <- hvlt_retention_long_who[-outlier_rows_hvlt_retention, ]
m1_hvlt_retention_clean <- lmer(score ~ group_who + timepoint_numeric + group_who*timepoint_numeric + age2 + sex_cat + years_of_ed + days_since_injury + group + (1|id), data = hvlt_retention_clean)
summary(m1_hvlt_retention_clean)

# Compare original vs cleaned model
compare_performance(m1_hvlt_retention_who, m1_hvlt_retention_clean)

# Recheck outliers on cleaned model
check_outliers(m1_hvlt_retention_clean)

# Assumptions check on final model
result_retention <- check_model(m1_hvlt_retention_clean)
plot(result_retention, panel = FALSE)

# standardized ES
standardize_parameters(m1_hvlt_retention_clean, method = "refit") %>%
  filter(Parameter == "group_who:timepoint_numeric")

means_df_hvlt_retention <- hvlt_retention_clean %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  ) %>%
  group_by(Group, TimePoint) %>%
  summarise(n = n(), Mean = mean(score, na.rm = TRUE), SE = sd(score, na.rm = TRUE) / sqrt(n), .groups = "drop")

individual_df_hvlt_retention <- hvlt_retention_clean %>%
  filter(!is.na(score)) %>%
  mutate(
    TimePoint = factor(timepoint_numeric, levels = c(0, 1), labels = c("Pre", "Post")),
    Group     = factor(group_who, levels = c(0, 1), labels = c("Did Not Meet", "Met Threshold"))
  )

ggplot() +
  geom_line(data = individual_df_hvlt_retention,
            aes(x = TimePoint, y = score, group = id, color = Group),
            alpha = 0.15, position = position_dodge(width = 0.2)) +
  geom_point(data = individual_df_hvlt_retention,
             aes(x = TimePoint, y = score, group = id, color = Group),
             alpha = 0.3, size = 2, position = position_dodge(width = 0.2)) +
  geom_line(data = means_df_hvlt_retention,
            aes(x = TimePoint, y = Mean, group = Group, color = Group),
            linewidth = 1.5) +
  geom_point(data = means_df_hvlt_retention,
             aes(x = TimePoint, y = Mean, group = Group, color = Group),
             size = 4) +
  geom_errorbar(data = means_df_hvlt_retention,
                aes(x = TimePoint, ymin = Mean - SE, ymax = Mean + SE, group = Group, color = Group),
                width = 0.1, linewidth = 1) +
  scale_color_manual(values = c("Did Not Meet" = "#40B0A6", "Met Threshold" = "#FFBE6A")) +
  labs(
    title    = "HVLT Retention (Higher scores are better)",
    subtitle = "Pre-Post Changes by Walking Threshold Group",
    x        = "Assessment Time",
    y        = "Score",
    color    = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

### Results Summary

```{r}
library(emmeans)

# ── helpers ──────────────────────────────────────────────────────────────────

fmt_p <- function(p) {
  if (p < .001) "< .001" else paste0("= .", sprintf("%03d", round(p * 1000)))
}

extract_stats <- function(label, model, int_term, group_var = "group_who") {

  ct  <- coef(summary(model))
  ci  <- confint(model, method = "Wald", parm = int_term)

  b   <- ct[int_term, "Estimate"]
  se  <- ct[int_term, "Std. Error"]
  df  <- ct[int_term, "df"]
  tv  <- ct[int_term, "t value"]
  cil <- ci[int_term, "2.5 %"]
  ciu <- ci[int_term, "97.5 %"]

  int_line <- sprintf(
    "  Interaction:      b = %.2f, SE = %.2f, t(%.2f) = %.2f, p %s, 95%% CI [%.2f, %.2f]",
    b, se, df, tv, fmt_p(p), cil, ciu)

  # within-group pre-to-post changes via emmeans
  emm_form <- reformulate(paste0(group_var, " * timepoint_numeric"))
  emm <- emmeans(model, emm_form, at = list(timepoint_numeric = c(0, 1)))
  wc  <- summary(contrast(emm, method = "revpairwise", by = group_var), infer = c(TRUE, TRUE))

  g0 <- wc[wc[[group_var]] == 0, ]
  g1 <- wc[wc[[group_var]] == 1, ]

  g0_line <- sprintf(
    "  Did Not Meet (0): b = %.2f, SE = %.2f, t(%.2f) = %.2f, p %s, 95%% CI [%.2f, %.2f]",
    g0$estimate, g0$SE, g0$df, g0$t.ratio, fmt_p(g0$p.value), g0$lower.CL, g0$upper.CL)

  g1_line <- sprintf(
    "  Met Threshold (1):b = %.2f, SE = %.2f, t(%.2f) = %.2f, p %s, 95%% CI [%.2f, %.2f]",
    g1$estimate, g1$SE, g1$df, g1$t.ratio, fmt_p(g1$p.value), g1$lower.CL, g1$upper.CL)

  cat(paste0("\n", label, "\n"))
  cat(int_line, "\n")
  cat(g0_line, "\n")
  cat(g1_line, "\n")
}

# ── run for each model ───────────────────────────────────────────────────────

extract_stats("PROMIS Cognitive Function",
              model_clean_2,
              int_term = "group_who:timepoint_numeric")

extract_stats("PROMIS Physical Health",
              m1_promis_phys,
              int_term = "group_who:timepoint_numeric")

extract_stats("PROMIS Mental Health",
              m1_promis_men,
              int_term = "group_who:timepoint_numeric")

extract_stats("Verbal Fluency",
              m1_vf_who,
              int_term = "group_who:timepoint_numeric")

extract_stats("HVLT Total Recall",
              m1_hvlt_total_who,
              int_term = "group_who:timepoint_numeric")

extract_stats("HVLT Delayed Recall",
              m1_hvlt_delayed_clean_3,
              int_term = "group_who:timepoint_numeric")

extract_stats("HVLT Retention",
              m1_hvlt_retention_who,
              int_term = "group_who:timepoint_numeric")
```

```{r}
library(kableExtra)

# ── helper: returns a single-row data frame instead of printing ───────────────
extract_stats_df <- function(label, model, int_term, group_var = "group_who") {

  ct  <- coef(summary(model))
  ci  <- confint(model, method = "Wald", parm = int_term)

  b   <- ct[int_term, "Estimate"]
  se  <- ct[int_term, "Std. Error"]
  df  <- ct[int_term, "df"]
  tv  <- ct[int_term, "t value"]
  cil <- ci[int_term, "2.5 %"]
  ciu <- ci[int_term, "97.5 %"]

  emm_form <- reformulate(paste0(group_var, " * timepoint_numeric"))
  emm <- emmeans(model, emm_form, at = list(timepoint_numeric = c(0, 1)))
  wc  <- summary(contrast(emm, method = "revpairwise", by = group_var), infer = c(TRUE, TRUE))

  g0 <- wc[wc[[group_var]] == 0, ]
  g1 <- wc[wc[[group_var]] == 1, ]

  data.frame(
    Outcome            = label,
    `Int. b (SE)`      = sprintf("%.2f (%.2f)", b, se),
    `t (df)`           = sprintf("%.2f (%.0f)", tv, df),
    `95% CI`           = sprintf("[%.2f, %.2f]", cil, ciu),
    `DNM b (SE)`       = sprintf("%.2f (%.2f)", g0$estimate, g0$SE),
    `MET b (SE)`       = sprintf("%.2f (%.2f)", g1$estimate, g1$SE),
    check.names        = FALSE,
    stringsAsFactors   = FALSE
  )
}

# ── build table ───────────────────────────────────────────────────────────────
results_table <- bind_rows(
  extract_stats_df("PROMIS Cognitive Function", model_clean_2,          "group_who:timepoint_numeric"),
  extract_stats_df("PROMIS Physical Health",    m1_promis_phys,          "group_who:timepoint_numeric"),
  extract_stats_df("PROMIS Mental Health",      m1_promis_men,           "group_who:timepoint_numeric"),
  extract_stats_df("Verbal Fluency",            m1_vf_who,               "group_who:timepoint_numeric"),
  extract_stats_df("HVLT Total Recall",         m1_hvlt_total_who,       "group_who:timepoint_numeric"),
  extract_stats_df("HVLT Delayed Recall",       m1_hvlt_delayed_clean_3, "group_who:timepoint_numeric"),
  extract_stats_df("HVLT Retention",            m1_hvlt_retention_who,   "group_who:timepoint_numeric")
)

# ── render ────────────────────────────────────────────────────────────────────
results_table %>%
  kbl(caption  = "Results Summary: Interaction and Within-Group Pre–Post Changes",
      booktabs = TRUE,
      align    = c("l", rep("c", 5))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                font_size  = 12) %>%
  add_header_above(c(" " = 1,
                     "Interaction (Group × Time)" = 3,
                     "Did Not Meet (pre→post)" = 1,
                     "Met Threshold (pre→post)" = 1)) %>%
  column_spec(1, bold = TRUE, width = "14em")
```